---
title: "Modelling the emergence of iconicity in interactive language learning"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2: default
  bookdown::html_document2: default
bibliography: latex-stuff/iconicity_lit.bib
---

```{r setup, include=FALSE}
library(knitr)

# Set knit defaults for code chunks
opts_chunk$set(
  dev = 'png', # default format of figures
  comment="",
  echo=FALSE, warning=TRUE, message=FALSE,
  cache=FALSE)

# some useful formatting functions for output of knitting
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

color_block = function(color) {
  function(x, options) sprintf('\\color{%s}\\begin{verbatim}%s\\end{verbatim}\\color{black}',
                               color, x)
}
knitr::knit_hooks$set(error = color_block('red'))
knitr::knit_hooks$set(warning = color_block('orange'))

base.width = 2.5
base.height = 2.5

```

```{r libraries, include=FALSE}
library(tidyverse)  # data wrangling and plotting
library(magrittr)   # for all pipes
library(ggplot2)    # for plotting
library(patchwork)  # combining plots
library(broom)      # for regression analysis
```

```{r functions, include=FALSE}
source("functions.R")
```

# TO DO

  * Check initialization values to be of roughly the same magnitude
  * Revisit the question of linear noise in a bounded system
  * Generate overview figure

```{r global-options, include=FALSE}
RESET_MODELS = F        # switch on/off whether model is re-run
```

# Introduction
<!--this entire section needs editing-->
Iconicity refers to the linguistic phenomenon in which the sound of a word resembles aspects of its meaning. In studies of language evolution, iconicity has often been regarded as playing only a marginal role. The conventionalized complexity of modern languages suggests that the relation between form and meaning is largely arbitrary, and even if iconicity could have been a driving force in the early emergence of language---facilitating the initial establishment of communicative conventions---it is expected to erode over time through semantic and phonological change. Yet, modern languages display examples of iconic forms that are both sustained and possibly continuously invented, indicating that iconicity can coexist with other types of structure, such as systematicity and combinatoriality [@johansson2020a; @johansson2020b; @sidhu2021; @winter2022]. From a learning perspective, it has been shown that iconic words are easier to learn, facilitating the mappings between label and signal [@imai-kita2014].

Empirically, the emergence of iconicity is often studied in experiments of artificial language learning, through iterated learning across generations of participants, or in language interaction games. The artificial language is represented by a set of form-meaning pairs, where the form is either a lexical item (sequences of graphemes), or an acoustic signal, such as a spoken lexical item (sequences of phonemes) or sound sequences (whistles, buzzer tones, etc.) [@hofer-levy2019; @kempe2019; @kempe2021; @verhoef2015; @verhoef2016]. The signal is assumed to convey an underlying semantic property of the item depicted in figures shown to participants. These figures differ along a perceptually salient semantic dimension referring to e.g., physical size (large vs. small), color, shape (spiky/pointy vs. round/fluffy), texture, motion (upward vs. downward movement), brightness (dark vs. light). In the mapping of an acoustic signal to a semantic referent, it has been shown that participants prefer high front vowels for smaller, brighter, and faster stimuli, and low back vowels for larger or rounded objects [see e.g., studies on the *bouba/kiki* effect; @ahlner-zlatev2010; @cwiek2022; @kohler1929]. Arguably, this association stems from the acoustic properties of vowels, where high front vowels resonate at frequencies that resemble those of small mammals [@ohala1983; @fitch1997]. While less is known about other acoustic properties of the speech signal, such as vowel duration, pitch trajectories, intensity patterns, studies have shown that participants tend to generate longer sounds for more complex objects [@vinson2021].

In the studies of artificial language learning, participants learn the form-meaning mappings and reproduce them multiple times, either in language interaction games, or across participant generations, where the reproductions of the first generation becomes the input for the second, etc. At reproduction, variability in pronunciation arises over trials, that can be affected by different biases. Arguably, iconicity is one such bias that can affect speakers to produce signals that resemble the underlying meaning of the referent. For instance, changing the signal to align iconically with the semantic property of the item by varying vowel quality, signal length, tone complexity, or number of segments in the signal.

Previous work has found iconicity to emerge during artificial language learning under certain circumstances. First, in settings where there is communicative pressure for successful interaction, such as a language learning game between two participants, iconic strategies are used and reused by participants [@kempe2019; @verhoef2015]. These can be further induced by rewarding participants for successful interactions (earning points, stars, etc.).

Second, as languages conventionalize, iconicity seems to loose its importance. Several studies have reported a trade-off between iconicity and the emergence of structured, combinatorial systems. For example, @hofer-levy2019 found that iconic mappings developed rapidly but gradually faded as combinatorial structure increased over successive interactions. Other work has shown that iconic forms can instead stabilize or even increase over repeated interactions [@perlman2015; @verhoef2015]. More broadly, studies of artificial language evolution demonstrate that languages become more learnable over time---through reductions in complexity and/or increases in combinatorial structure---while iconicity either stabilizes or declines [@hofer-levy2019; @vinson2021]. Verhoef and colleagues (2016) further observed that iconic forms persisted longer under conditions where opportunities for structural regularization were delayed, such as when form–meaning mappings were scrambled across generations.

Third, research has shown that shorter training tends to favor iconic forms: when learners cannot rely on rote memorization of form–meaning mappings, they make greater use of iconicity [@kempe2021]. In @kempe2019, limited training and the absence of explicit communicative demands led to degeneration and reduced distinctiveness, with neither iconicity nor compositional structure emerging. Iconicity and some structural regularities only appeared when participants engaged in a referential communication task. In @kempe2021, combinatorial structure emerged under both long and short training conditions, but iconicity was specifically strengthened when learning opportunities were limited, with learners exploiting form–meaning resemblance (e.g., shorter signals for small referents).

Finally, some studies have also indicated differences in the strength of iconicity as dependent on the semantic dimension, with stronger effects for size over shape [@johansson2021]. Comparing spoken and written domain, iconicity has been found to faster emerge in spoken domains [@vinson2021]. 

To sum up, iconicity tends to emerge when learners cannot rely on rote memorization, when meaning (the semantic property of the item) is salient, and when communicative pressure is present. Combinatorial patterns tend to increase with repetition and transmission but are sometimes inhibited by highly iconic or semantically transparent mappings. Learnability typically improves over generations, but whether this improvement is driven by structure or iconicity depends on the design of each study.

# Suggested mechanisms underlying the emergence of iconicity
Based on what we know from previous work, we suggest three main mechanisms supporting the emergence of iconicity in interactive language learning. Based in the idea that iconicity facilitates interaction and learning of an artificial language, these mechanisms target different parts of the interaction, affecting either the signal being produced, the interpretation ease of the signal during recognition, or the memory trace of the learned signal after recognition. The three mechanisms encompass: an *adaptive production bias* toward producing more iconic signals based on previous interactional outcomes, a *interpretation bias* that favor iconic signals and punishes anti-iconic signals in recognition, and finally, an *adaptive signal alignment* toward making learned signals more iconic when communication fails.

## Adaptive production bias
In an interaction game, iconic signal formation can be seen as a bias in adaptive production whereby speakers adjust their signal to be more iconic depending on previous communication outcomes. Given that iconicity strategies arise from communicative pressure, one can expect this production bias toward a more iconic signal to be larger if previous communication was unsuccessful. This would reflect a situation where interlocutors have difficulties communicating in the new language, and the speaker would resort to using iconicity as a means for increasing the probability of successful interaction. The *adaptive production bias* mechanism thus represents a directional pressure toward iconic targets during production, causing iconicity to gradually emerge over repeated interaction.

## Interpretation bias
If iconic signals indeed drive interaction and learning, iconic forms should further be favored in recognition. Similarly, non-iconic forms should be dis-favored. The *interpretation bias* mechanism incentivizes iconic signals based on their distance to prototype; boosting the ease of interpretation for signals that are closer to the prototype, and punishing signals that are further away from the prototype. This means that the listener's success in identifying the referent increases with signal proximity to prototype, and iconicity thus boosts perception and interpretability. In a situation where the listener tries to identify what semantic referent the speaker is signalling, more iconic signals would thus be easier for the listener to identify than non-iconic signals. This means that iconic forms become preferred because they are easier for listeners to interpret and therefore more likely to be reinforced. 

## Adaptive signal alignment
In interaction, feedback on communicative success is expected to have an effect on language learning and signal representation in memory, where both successful and unsuccessful interactions can shape representations. If iconicity is used as a strategy for communicative success, one could expect an intentional pressure on listeners to update their learned signals in line with interactional outcomes. On failed interactions, the signals are shifted toward the prototype, thus becoming more iconic, and less so on successful interactions. The *adaptive signal alignment* mechanism could be interpreted as an intentional effort on behalf of the listener to shift the memory representation of the signal toward the prototype when communication fails, aiming to make the form-meaning mapping easier to recognize in upcoming interactions. This means that iconicity does not only affect the production of signals, but also the memory of the signal-referent mapping.

Next, we describe how these suggested mechanisms are implemented in a computational model simulating the emergence of iconicity.

# Model implementation and underlying assumptions
<!--There have been few attempts at computationally model the emergence of iconicity in interaction games [for a review, see @dona-schouwstra2024]. The studies have often used agent-based learning models...-->

The model we introduce here simulates the emergence of iconicity in an interactive game of artificial language learning, played in dyads. Guided by communicative pressure and incremental learning, the above mechanisms are implemented in the model as parameters supporting the emergence of iconicity. The central assumption is that successful interaction favors form-meaning mappings that are iconic, and that over repeated interactions, iconic signals emerge. Learning is considered gradual and experience-dependent as agents continuously update their internal representations of form-meaning associations and adjust their signal production based on interaction outcomes (success or failure). Success or failure is formalized as the probability of the current guess, $p$. Because learning is considered probabilistic, increases in guessing probabilities do not guarantee perfect performance on every trial, but raise the probability of correct recognition.

The emergence pressure of iconicity is formalized in two sets of parameters, in line with the above described mechanisms: one set of parameters operating on the listener's perception and retrieval of the referent, and on the storage of the signal after recognition, and one set of parameters operating on the production of the signal. We begin by accounting for the overall structure of the model, before describing how these parameters are implemented.

## Overall structure of model
Any simple agent-based model of lexical learning that simulates the emergence of iconicity needs to account for 1) agents and the language to be learnt, 2) how agents produce and perceive form-meaning pairs, 3) how learning incrementally increases over time, and 4) the iconicity mechanisms affecting this process. As accounted for above, for iconicity to emerge, mechanisms need to exert pressure on agents to favor more iconic signals in production and perception, and in signal representation. The output of interest is whether the signals becomes more iconic over time. In this model, this is assessed using a distance measure of the signal to the prototypical value in a continuous signal space. More iconic signals are closer located to the prototypical value of the underlying referent. This goes for both learned and produced signals.

In the language interaction game simulated here, size-related meanings are conveyed using acoustic signals. There are a limited number of referents that need to be learned and communicated. Half of the referents are noticeably smaller than the other half of the referents. Physical size is assumed to be perceptually salient, consistent with extensive evidence that magnitude distinctions are readily mapped onto acoustic dimensions. Each referent has their own idiomatic meaning which is assumed to be unambiguously signaled by sequences of consonants that are presented to the agents alongside the semantic property of the referent. Whether the referent belongs to the large or small subset is communicated by the acoustic signal space. The acoustic signal thus does not actually signal lexical meaning but the underlying semantic aspect, meaning that it is not necessary for lexical retrieval; it is merely an additional source of information for perception of the underlying semantic aspect. This mirrors real perception where one can mis-hear a phonetic detail but still perceive the underlying size dimension. Since lexical retrieval can therefore be achieved without relying on iconicity biases per se, iconicity is treated as facilitating recognition rather than being necessary for it. This motivates a central assumption in this model of constant increase in learning over interaction rounds, on both successful and unsuccessful trials. Agents can thus learn the mappings even without depending on the iconic signal, though iconicity may aid communication. In addition, it means that the guessing probability refers to the semantic property of the item, and not to the lexical item itself (the identification of the lexical item is not modeled here).

The model loops through several trials in the experiment where two participants, agent A and agent B, take turns alternating roles as listener and speaker. Each trial is initialized by randomly assigning agent A and agent B to either speaker or listener role. The speaker is given a random referent (small or large) and the learned signal for that referent is retrieved. The speaker then produces a signal, and the listener attempts to recognize the intended referent. The recognition is evaluated as either correct or incorrect, and both speaker and listener update their beliefs (= their learning) about the signal-to-referent mapping independently of iconicity, continuously becoming more and more certain about the language over trials. Finally, the listener updates its learned signal representation depending on the interactional outcome, and a new trial starts.

## Implementation of suggested mechanisms
The suggested mechanisms are implemented at different stages of the interaction round accounted for above. Figure XX visualizes a single trial and highlights at what stages the mechanisms are assumed to operate.

### Producing a signal
Beginning at the bottom of Figure XX, the speaker agent is given a referent at random and retrieves the stored signal values [x,y] for the referent. Signals are represented as points in a continuous (yet contained) two-dimensional signal space ranging from [0,0] to [1,1]. This space can be considered a simplified abstraction over multiple correlated acoustic dimensions (e.g., vowel quality, duration, pitch) that jointly encode size-related meaning. The two dimensions do not correspond to specific acoustic parameters but capture the idea that signals vary continuously along multi-dimensional perceptual dimensions relevant for iconic mapping. Each meaning is associated with an iconic prototype in signal space. Small referents are associated with a prototype at [0,0], and the large category with a prototype at [1,1]. These prototypes represent perceptually transparent signal–meaning correspondences and can be thought of e.g. a formant vowel space in which high front vowels are correlated with small referents and low back vowels are correlated with large referents. The initial signals at the start of the simulation are placed at the center of the space, [.5,.5].

When the speaker retrieves the signal for the randomly selected referent, the *adaptive production bias* mechanism transforms the signal depending on the speaker's current probability of correct identification, $p$. The initial learning state of the listener at the first trial is on average a probability of $~.33$. This probability is established by drawing n random values from a beta distribution with a $\alpha = 2$ and $\beta = 4$. The updated biased signal is computed as a weighted average of the learned signal and the prototype, where the prototype weight increases as $p$ decreases, so as to be modulated by the speaker's confidence in the learned form-meaning mappings; lower current guess probabilities generate stronger prototype bias, whereas higher guess probabilities generate weaker bias ($1-p_{speaker}$). The adaptive production bias is implemented as sampling from a two-dimensional Gaussian probability distribution with a mean shifted toward [0,0] or [1,1]. Signals for small referents tend to shift toward the [0,0] prototype while signals for large referents tend to shift toward the [1,1] prototype.

After updating the produced signal, *Gaussian noise* is applied to each dimension of the produced token with $sd = .05$ to model production variability (incl. e.g., motor noise). After noise is applied, values are bounded within [0,1] in each dimension. The production noise affects only the produced token and does not modify the underlying stored representation.

```{r echo=TRUE, eval=FALSE}
produce_signal <- function(previous_signal, referent_type, speaker_guess_prob) {
  target <- prototype_point(referent_type)
  bias_strength <- adaptive_production_bias * (1 - speaker_guess_prob)
  mu <- (1 - bias_strength) * previous_signal + bias_strength * target
  clamp01(rnorm(2, mean = mu, sd = drift_sd))
}
```

### Recognizing the signal
Moving up in Figure XX, the biased signal is then presented to the listener, that guesses what referent the signal refers to. Recognition and learning of the form-meaning mappings are formalized as a function of the listener's previous guessing probability and the *interpretation bias* term. Guessing probability reflects the listener's prior probability of guessing a specific referent correctly and corresponds to the associative strength linking a signal to a meaning. Higher previous guessing probabilities generate higher probability of success on subsequent trials, thus reinforcing existing mappings and accelerating learning trajectories. 

The interpretation bias is implemented as the signal's distance to the iconic prototype affecting the current guess, favoring signals that are more closely located to their prototypical target in the two-dimensional space, and punishing signals that are further away from target. It is operationalized by two parameters---*signal evidence* and *prototype weighting*. Signal evidence measures the Euclidean distance between a produced signal and its iconic prototype, capturing how well the signal fits the attractor. It is defined as a decreasing function that drops off sharply (exponentially) when moving away from prototypes. Figure \@ref(fig:signal-ease) illustrates signal evidence relative to the signal space, with higher ease for more prototypical values.

(ref:signal-ease) Illustration of signal evidence in the two-dimensional signal space, for small and large referents.

```{r signal-ease, fig.width=base.width*4, fig.height=base.height*2, out.width='90%', fig.align='center', fig.pos='ht', fig.cap="(ref:signal-ease)"}
# show heatmap of ease relative to signal space
grid <- expand.grid(
  x = seq(0, 1, length.out = 50),
  y = seq(0, 1, length.out = 50),
  type = c("small", "large"))

# make heat map
grid <- grid %>%
  rowwise() %>%
  mutate(evidence = signal_evidence(c(x, y), type)) %>%
  ungroup()

ggplot(grid,
       aes(x = x, y = y, fill = evidence)) +
  geom_tile() +
  facet_wrap(. ~type) +
  scale_fill_viridis_c()

```

The signal evidence is scaled by the prototype weighting parameter which modulates the impact of signal distance affecting guessing. Iconic signals are scaled by the prototype weighting parameter, while non-iconic signals are scaled by the same parameter with opposite sign. This reflects a mechanism that favors iconic signals and punishes anti-iconic signals in recognition. Together, these parameters make form-meaning pairs easier or more difficult to perceive, encode, and retrieve, depending on their spatial proximity to iconic prototypes.

```{r echo=TRUE, eval=FALSE}
signal_evidence <- function(signal, type, distance_scale = sqrt(2)) {
  target <- prototype_point(type)
  dist <- sqrt((signal[1] - target[1])^2 + (signal[2] - target[2])^2)
  evidence <- 1 - (dist / distance_scale) * 2
  evidence
}
```

Guessing involves uncertainty which leads to imperfect learning. This uncertainty reflects both perceptual noise (variability in processing the acoustic signal) and attentional lapses (uninformed responses that ignore the signal). The guessing probability is computed in log-odds space by combining the listener's prior guessing probability with the iconicity-weighted signal evidence, and then converting back to probability space. A soft lapse mechanism is then applied, bounding the resulting probability at $p_{max} = .95$, corresponding to a $.05$ lapse rate in which responses are uniform rather than signal-driven. This prevents probabilities from collapsing to 0 or 1 and models the fact that agents are not perfectly attentive or rational on every trial.

The final outcome of the trial, success or failure, is implemented as random sampling from a binomial distribution with probability $p$ = current guessing probability. A value of 1 indicates that the referent was correctly recognized, while 0 indicates an incorrect guess. 

After calculating success or failure, learning is reinforced for both speaker and listener by implementing the *associative learning bias* mechanism. This is a learning mechanism and not an iconicity mechanism. It is parameterized as applying a baseline learning strength to the speaker's and listener's guessing probabilities, scaled slightly depending on the recognition outcome: successful interactions lead to a moderately larger increase in guessing probability, while unsuccessful interactions produce a slightly smaller update. Specifically, the update on correct trials is scaled by a factor of $1.2$, further reinforcing learning, and the update on incorrect trials is scaled by $.8$, keeping the overall magnitude of learning roughly consistent across trials. This reflects the assumption that errors and feedback incrementally strengthen associations over time, with stronger learning on correct interactions.
According to previous work on general learning, negative feedback has a stronger influence on subsequent interactions...

```{r echo=TRUE, eval=FALSE}
guess_probability <- function(agent_guess_prob, signal_xy, referent_type) {
  evidence <- signal_evidence(signal_xy, referent_type)
  p_logit <- qlogis(agent_guess_prob) + prototype_weight * evidence
  p_raw <- plogis(p_logit)
  p_final <- (lapse/2) + (1 - lapse) * p_raw
  p_final
}

update_guess_logodds <- function(p_old, delta) {
  plogis(qlogis(p_old) + delta)
}

p_correct <- guess_probability(listener_guess[r], sig_prod, r_type)
success <- rbinom(1, 1, p_correct)
delta <- if (success == 1) {
  2 * learning_strength
} else {
  0.5 * learning_strength
}

speaker_guess[r] <- update_guess_logodds(speaker_guess[r], delta)
listener_guess[r] <- update_guess_logodds(listener_guess[r], delta)
```

### Updating the learned signal
After the listener and speaker have received feedback on the interaction, they update their learned signals in memory, Figure XX:top. The updating is affected by the *adaptive signal alignment* mechanism that aligns the signal toward either the produced or the prototypical signal depending on whether interaction succeed or failed. The adaptive signal alignment is implemented by two parameters: a *reinforcement rate* on successful interactions, and a *corrective rate* on failed interactions. 

When communication succeeds, the listener retrieves the signal for the referent and the learned signal is updated toward the speaker's produced signal by a fixed proportion (reinforcement rate) in probability space. When communication fails, the learned signal is instead moved toward the iconic prototype for that referent by a fixed proportion (corrective rate). In addition, noise is added to the failed signal to capture variability in updating failed signals.

```{r echo=TRUE, eval=FALSE}
reinforce_signal <- function(previous_signal, produced_signal) {
  clamp01((1 - reinforcement_rate) * previous_signal + reinforcement_rate * produced_signal)
}

update_signal_on_failure <- function(previous_signal, referent_type) {
  target <- prototype_point(referent_type)
  clamp01(previous_signal + corrective_rate * (target - previous_signal) + rnorm(2, 0, drift_sd/2))
}

new_sig <- if(success == 1) {
  reinforce_signal(previous_signal, sig_prod)
} else {
  update_signal_on_failure(previous_signal, r_type)
}
signals$x[r] <- new_sig[1]
signals$y[r] <- new_sig[2]
```

Finally, the speaker and listener guessing probabilities are updated based on their current probabilities, before the output is stored and the agents proceed to the next trial.

Next, we compare the different mechanisms and assess their relative importance for the emergence of iconicity.

```{r}
```

# Comparison of mechanisms and their relative importance for the emergence of iconicity
For the comparison presented below, we ran 10 simulations of the model, each simulation with 1000 interaction rounds for a total of 10 000 trials. Simulations were run with fixed random seeds for reproducibility.<!--changing the seed did not qualitatively alter the results -- haven't checked if that is actually true--> The number of unique referents in the guessing game was set to 6; 3 small referents and 3 large referents.

The simulations used the following parameter settings: learning strength ($.005$), prototype weighting ($.2$), adaptive production bias ($.15$), reinforcement rate ($.05$), and corrective rate ($.03$).<!--theses values should have approx. the same magnitude, depending on their role in respective function-->

We generated 6 types of models. The main model included all of the above described parameters set at their intialization values, as described above. The remaining 5 models were generated by switching off one of the parameters at a time by setting their value to 0.

Figure \@ref(fig:iconicity-plot) illustrates the emergence of iconicity over trials for each of these 6 models. All models display similar patterns of iconicity emergence over time, besides the model for which failure drift is switched off. This highlights the importance of pushing produced signals on failed interactions toward more prototypical values. Iconicity might thus not be a by-product of communicative success, but requires specific signal-space dynamics. Interestingly, the best performing model seems to be the model where no learning strength is added to the updated speaker and listener guess rates.

(ref:iconicity-plot) Evolution of iconicity for learned tokens across simulation time. Iconicity is averaged in bins of 100 rounds and showed separately for each parameter setting (facets). Light lines trace individual simulations, the bold line denotes the average across simulations. The dashed horizontal line represents baseline iconicity expected from random signals without structure or learning.

```{r parameters}
set.seed(2139752)

# Generate empty dataframe for simulation
d.empty <- data.frame(
  sim      = integer(),
  round    = integer(),
  referent = integer(),
  speaker  = character(),
  listener = character(),
  type     = character(),
  p_correct = numeric(),
  success   = integer(),
  stored_x  = numeric(),
  stored_y  = numeric(),
  produced_x = numeric(),
  produced_y = numeric(),
  stringsAsFactors = FALSE)

# Generate one data frame with different simulations where params are switched on/off
d.simulation <- rbind(
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000) %>%
    mutate(parameter_space = "allParams"),
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, adaptive_production_bias = 0) %>%
    mutate(parameter_space = "noAdaptiveProductionBias"),
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, corrective_rate = 0) %>%
    mutate(parameter_space = "noCorrectiveRate"),
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, reinforcement_rate = 0) %>%
    mutate(parameter_space = "noReinforcement"),
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, learning_strength = 0) %>%
    mutate(parameter_space = "noLearning"),
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, prototype_weight = 0) %>%
    mutate(parameter_space = "noPrototypeWeight"),
  # Also add one simulation that resembles the old model
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, adaptive_production_bias = 0, corrective_rate = 0, reinforcement_rate = 0) %>%
    mutate(parameter_space = "onlyLearning"))

# Diagnostics
# d.onlyDrift <- d.empty %>%
#   run_interaction_sim(n_sim = 10, n_rounds = 1000, learn_rate = 0, prod_bias = 0, reinforcement_rate = 0, iconicity_boost = 0) %>%
#   mutate(parameter_space = "onlyDrift")

```

```{r iconicity}
# you can compute iconicity either from stored signals or produced tokens
d.iconicity <- d.simulation %>%
  group_by(parameter_space) %>%
  mutate(bins = cut(round, breaks = 20, labels = FALSE)) %>%
  group_by(parameter_space, bins, type, sim) %>%
  summarise(
    x_learned = mean(stored_x),
    y_learned = mean(stored_y),
    x_produced = mean(produced_x),
    y_produced = mean(produced_y),
    .groups = "drop") %>%
  pivot_longer(
    cols = starts_with(c("x", "y")),
    names_to = c("signal", "process"),
    values_to = "value",
    names_sep = "_") %>%
  pivot_wider(
    names_from = signal,
    values_from = value) %>%
  mutate(
    target_x = ifelse(type == "small", 0, 1),
    target_y = ifelse(type == "small", 0, 1),
    dist = sqrt((x - target_x)^2 + (y - target_y)^2),
    iconicity = exp(-2 * dist)) %>%
  group_by(parameter_space, bins, sim, process) %>%
    summarise(iconicity = mean(iconicity), .groups = "drop")

```

```{r iconicity-plot, fig.width=base.width*6, fig.height=base.height*3, fig.align='center', out.width='100%', fig.cap="(ref:iconicity-plot)", warning=FALSE}
#iconicity trajectories - remove model with only learning since that was added to compare against old model
p.learned.iconicity <- d.iconicity %>%
  filter(process == "learned", parameter_space != "onlyLearning") %>%
  group_by(parameter_space, bins, sim) %>% 
  summarise(iconicity = mean(iconicity)) %>%
  ggplot(
    aes(x = bins, y = iconicity, group = sim)) +
  geom_path(size = 0.9, alpha = 0.2) +
  geom_path(data = . %>% 
              group_by(parameter_space, bins) %>%
              summarise(iconicity = mean(iconicity)), 
            aes(group = 1), size = 1.5) +
  geom_hline(yintercept = exp(-2 * sqrt((0.5)^2 + (0.5)^2)), lty = "dashed") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
  labs(title = "Iconicity over time (learned tokens)",
       y = "Iconicity (high = iconic)", x = "Bins (rounds in 100s)") +
  theme_minimal() +
  facet_grid(~parameter_space)

# p.produced.iconicity <- p.learned.iconicity %+%
#   (d.iconicity %>%
#   filter(process == "produced") %>%
#   group_by(parameter_space, bins, sim) %>% 
#   summarise(iconicity = mean(iconicity))) +
#   labs(title = "Iconicity over time (produced signals)") +
#   facet_grid(~parameter_space)

p.learned.iconicity

# d.simulation %>%
#   mutate(
#     dx = produced_x - stored_x,
#     dy = produced_y - stored_y) %>%
#   group_by(parameter_space, success) %>%
#   summarise(
#     mean_step = mean(sqrt(dx^2 + dy^2)),
#     .groups = "drop")

```

We also evaluated the models in terms of the overall success rate across time, that is, the communicative accuracy of agents in interaction. The accuracy for each of the 6 models is strikingly similar, except for the model where no successful learning bias is applied, see Figure \@ref(fig:accuracy). This seems to suggest that guessing accuracy depends entirely on learning bias (which is expected given that it is the only mechanism that updates guessing probabilities).

(ref:accuracy) Success rate across simulation rounds. A binomial GLM with smoothing is fit to the proportion of successful interactions. Shaded bands indicate the model-based standard error. Panels correspond to the different parameter settings.

```{r accuracy, fig.width=base.width*6, fig.height=base.height*3, fig.align='center', out.width='100%', fig.cap="(ref:accuracy)"}
# Plot success rate over time across simulation within parameter space
d.simulation %>%
  filter(parameter_space != "onlyLearning") %>%
  #group_by(parameter_space, round) %>%
  #summarise(success = mean(success)) %>%
  ggplot(
    aes(x = round, y = success)) +
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"), 
              se = TRUE) +
  labs(title = "Success rate over time (across simulations)", 
       y = "Success", 
       x = "Round") +
  facet_grid( ~ parameter_space) +
  scale_y_continuous(limits = c(0,1)) + 
  theme_minimal()
```

Figure \@ref(fig:signal-space) visualizes the distribution of learned and produced signals across simulations for each of the 6 model types. The figure seems to suggest that categories separate even for the model where iconicity does not emerge strongly, e.g., the model without corrective rate (second row).

(ref:signal-space) Distribution of stored (left panels) and produced (right panels) signals over the course of the simulation. Each point represents a signal token in two-dimensional signal space; colors denote referent type (small vs. large). Columns correspond to simulation replicates and rows to parameter settings.

```{r signal-space, fig.width=base.width*7, fig.height=base.height*4, fig.align='center', out.width='100%', fig.cap="(ref:signal-space)"}
# Signal space use
p.learned.signals <- d.simulation %>%
  filter(parameter_space != "onlyLearning") %>%
  ggplot(
    aes(stored_x, 
        stored_y, 
        color = type)) +
  geom_point(alpha = 0.2) +
  labs(title = "Evolution of learned signal space", x = "x", y = "y") +
  facet_grid(parameter_space ~ sim) +
  scale_x_continuous(limits = c(0,1)) + 
  scale_y_continuous(limits = c(0,1)) + 
  scale_color_viridis_d(begin = 0.1, end = 0.9) +
  theme_minimal()

p.produced.signals <- p.learned.signals %+%
  aes(produced_x,
      produced_y) +
  labs(title = "Evolution of produced signal space", x = "x", y = "y")

p.learned.signals + p.produced.signals +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "bottom")

```

Overall, the results seem to suggest that iconicity emerged reliably except when the corrective rate mechanism was disabled. Removing failure drift eliminated the tendency for signals to move toward their iconic prototypes on failed trials, indicating that this mechanism plays a central role in shaping the signal space. This suggests that iconicity in this model depends on specific signal-space dynamics that bias failed forms toward iconic attractors.

Interestingly, learning strength appeared to influence communicative accuracy without substantially affecting iconicity. When learning strength was removed, agents reached relatively high levels of iconicity but reduced communicative success, indicating that learning strength governs guessing accuracy whereas failure drift governs the emergence of form-meaning structure. This seems to suggest that agents can learn to communicative successfully without producing iconic forms, and that iconic forms can emerge without directly improving recognition accuracy.

Visualizing the learned and produced signal spaces further suggested that category separation can occur even in the absence of iconicity, indicating that lexical categories can form without alignment to iconic prototypes. It might be the case that iconicity emerges from indirect learning advantages rather than from explicit pressure for communicative success---the combination of failure drift, adaptive production bias, interpretative ease, and reinforcement creating a positive feedback loop where failure pushes signals toward iconic prototypes, iconic signals are easier to interpret, successful interpretations reinforce produced forms and reinforcement consolidates iconic patterns in memory?<!--is it?--->

## Grid search over the parameter space

In order to investigate what parameter settings generate the largest increase in iconicity across trials and simulations, we generated a grid search over the five model parameters: adaptive production bias, prototype weighting, reinforcement rate, learning rate, and corrective rate. The grid search explored the following values for each of the parameters.

```{r echo=TRUE, eval=FALSE}
adaptive production bias = seq(0, 0.3, length.out = 4)
prototype weight = seq(0, 0.4, length.out = 4)
reinforcement rate = seq(0, 0.1, length.out = 4)
corrective rate = seq(0, 0.06, length.out = 4)
learning strength = seq(0, 0.1, length.out = 4)
```

Similarly to the figures presented above, each panel of Figure \@ref(fig:grid-search) visualizes iconicity across the parameter space for when one of the parameters is switched off. Replicating previous figures, Figure \@ref(fig:grid-search) shows that 1) iconicity struggles to emerge when corrective rate is switched off, 2) learning strength does not seem to support the emergence of iconicity, and 3) switching off the other three parameters (reinforcement rate, adaptive production bias, prototype weighting) generates highly similar patterns of iconicity, suggesting that these might play less of a role in the model.

For most combinations of parameters, the trade-off between learning strength and iconicity is evident---iconicity is higher for when learning strength = 0, suggesting that the more I learn for free, the less iconic I become (upper two panels, and bottom-right panel, Figure \@ref(fig:grid-search)). This would suggest that iconicity might not continue to grow in a language, at least not as a learning mechanism, but perhaps as a strategy to make the language more recognizable for the interlocutor.

We also note that under some conditions, iconicity still emerges when corrective rate is switched off (bottom-left panel, Figure \@ref(fig:grid-search)). This suggests that failure-based drift is not strictly necessary but is normally the dominant pathway (error-driven vs. success-driven iconicity). If this mechanism is not available, iconicity can still emerge by generating a positive feedback-loop of biased production--higher success--reinforcement--more biased learned signals--higher success.

(ref:grid-search) Heatmaps of iconicity across the parameter space for different model configurations. Each panel shows simulations where one parameter is disabled: no reinforcement, no adaptive production bias, no signal drift, no learning, or no prototype weighting. Color indicates mean iconicity; yellow outlines mark the iconicity baseline (\~.25) and red crosses indicate the parameter values that generates maximum iconicity.

```{r}
if(RESET_MODELS || !file.exists("../models/grid-search.rds")) {
  
  # Generate a parameter grid to explore
  d.param_grid <- expand.grid(
    adaptive_production_bias = seq(0, 0.3, length.out = 4),
    prototype_weight = seq(0, 0.4, length.out = 4),
    reinforcement_rate = seq(0, 0.1, length.out = 4),
    corrective_rate = seq(0, 0.06, length.out = 4),
    learning_strength = seq(0, 0.1, length.out = 4))
  
  # Prepare results container
  d.grid_results <- d.param_grid %>%
    mutate(iconicity = NA_real_,
           history = vector("list", n()))
  
  for (i in seq_len(nrow(d.param_grid))) {
    
    params <- d.param_grid[i, ]
    
    # empty structure for simulation log
    empty_df <- data.frame(
      sim = integer(), round = integer(), referent = integer(),
      speaker = character(), listener = character(), type = character(),
      p_correct = numeric(), success = integer(),
      stored_x = numeric(), stored_y = numeric(),
      produced_x = numeric(), produced_y = numeric(),
      stringsAsFactors = FALSE
    )
    
    # run simulation
    history <- run_interaction_sim(
      data = empty_df,
      n_sim = 10,
      n_referents = 6,
      n_rounds = 1000,
      drift_sd = 0.05,
      learning_strength = params$learning_strength,
      prototype_weight = params$prototype_weight,
      adaptive_production_bias = params$adaptive_production_bias,
      reinforcement_rate = params$reinforcement_rate,
      corrective_rate = params$corrective_rate,
      lapse = 0.05)
    
    # store full dataframe
    d.grid_results$history[[i]] <- history
    
    # compute iconicity score
    d.grid_results$iconicity[i] <- compute_iconicity(history)
    
    message("Completed parameter set ", i, " of ", nrow(d.param_grid))
  }
  
  
  saveRDS(d.grid_results, file = "../models/grid-search.rds", compress = T)
  
} else { d.grid_results <- readRDS("../models/grid-search.rds") }

d.full.history <- bind_rows(d.grid_results$history, .id = "param_set")

```

```{r grid-search, fig.width=base.width*6, fig.height=base.height*4, fig.align='center', out.width='100%', fig.cap="(ref:grid-search)"}
# plot for different param settings
p.noReinforce <- d.grid_results %>%
  mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
  filter(reinforcement_rate == 0) %>%
  #filter(corrective_rate == 0) %>%
  rename(`corrective rate` = corrective_rate, `adaptive production bias` = prod_bias) %>%
  ggplot(
    aes(
      x = learn_rate, 
      y = iconicity_boost, 
      fill = iconicity,
      color = ifelse(iconicity != max(iconicity), "red", "black"))) +
  geom_tile() +
  geom_tile(
    data = . %>%
      filter(iconicity >= 0.24 & iconicity <= 0.26),
    color = "yellow",
    size = .3,
    fill = NA) +
  geom_point(
    data = . %>%
      filter(iconicity == max(iconicity)),
    shape = 4) +
  scale_color_manual(values = c("red", "black")) +
  scale_fill_viridis_c() +
  facet_grid(`adaptive production bias` ~ `corrective rate`, labeller = label_both) +
  theme_minimal() +
  guides(color = "none") +
  labs(
    title = "Iconicity across the parameter space (no reinforcement)",
    x = "Learning rate",
    y = "Prototype weighting",
    fill = "Iconicity")

p.noProdBias <- p.noReinforce %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(prod_bias == 0) %>%
     rename(`corrective rate` = corrective_rate, reinforcement = reinforcement_rate)) +
  facet_grid(reinforcement ~ `corrective rate`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space (no adaptive production bias)")

p.noDrift <- p.noReinforce %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(corrective_rate == 0) %>%
     rename(reinforcement = reinforcement_rate, `adaptive production bias` = prod_bias)) +
  facet_grid(reinforcement ~ `adaptive production bias`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space (no corrective rate)") +
  guides(fill = "none")

p.NoLearning <- p.noReinforce %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(learn_rate == 0) %>%
     rename(reinforcement = reinforcement_rate, `corrective rate` = corrective_rate)) +
  aes(x = prod_bias) +
  facet_grid(reinforcement ~ `corrective rate`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space (no learning strength)") +
  scale_x_continuous("Adaptive production bias") +
  guides(fill = "none")

p.NoBoost <- p.noReinforce %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(iconicity_boost == 0) %>%
     rename(`reinforcement` = reinforcement_rate, `corrective rate` = corrective_rate)) +
  aes(y = prod_bias) +
  facet_grid(reinforcement ~ `corrective rate`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space (no prototype weighting)") +
  scale_y_continuous("Adaptive production bias") +
  guides(fill = "none")

(p.noReinforce  +
  guides(fill = "none") + p.noProdBias + plot_spacer()) /
  (p.noDrift + p.NoLearning + p.NoBoost) +
  plot_layout(widths = c(1,1,1),
              heights = c(1,1))
```

```{r}

```

# Open questions

-   learning update: the guess probabilities improve over time while corrective rate decreases (no drift on correct trials), this could mean that the signal tends to freeze near its initial state [.5,.5]?
-   Why the accuracy-iconicity dissociation?
-   Why the emergence of a binary category space even for models where iconicity does not emerge? Could it be that separation is emerging from Bayesian updating + task structure? informative feedback structure?

# References




