---
title: "Simple iconicity model"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: latex-stuff/iconicity_lit.bib
---

```{r setup, include=FALSE}
library(knitr)

# Set knit defaults for code chunks
opts_chunk$set(
  dev = 'png', # default format of figures
  comment="",
  echo=FALSE, warning=TRUE, message=FALSE,
  cache=FALSE)

# some useful formatting functions for output of knitting
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

color_block = function(color) {
  function(x, options) sprintf('\\color{%s}\\begin{verbatim}%s\\end{verbatim}\\color{black}',
                               color, x)
}
knitr::knit_hooks$set(error = color_block('red'))
knitr::knit_hooks$set(warning = color_block('orange'))

base.width = 2.5
base.height = 2.5

```

```{r libraries, include=FALSE}
library(tidyverse)  # data wrangling and plotting
library(magrittr)   # for all pipes
library(ggplot2)    # for plotting
library(patchwork)  # combining plots
library(broom)      # for regression analysis
```

```{r functions, include=FALSE}
source("functions.R")
```

```{r global-options, include=FALSE}
RESET_MODELS = F        # switch on/off whether model is re-run
```

# Modelling iconicity
In this document, we report on a simple agent-based model of the emergence of iconicity in an artificial language interaction game, played in dyads. First, we briefly describe the design of the experiment simulated by the model. Second, we describe the model, its parameters and their psychological grounding. Finally, we present results from simulations.

## Experiment design
In the experiment, two participants play an interactive game of artificial language learning. The language consists of XX form-meaning pairs that differ in size, shape and color. Each target referent is represented by a picture of an alien and a fixed consonant context (C_C_C) with inter-syllabic schwa-vowels. While the consonant context for each referent is always fixed, the vowel sounds can be altered by participants. 

Prior to game play, both participants, agent A and agent B, are trained on the XX form-meaning pairs up to a certain threshold. When both agents have met the learning criteria, they proceed to the interaction phase. On each trial in the interaction phase, agents take turns alternating roles of listener and speaker. The speaker agent gets a random referent (picture and consonant context) and produces a signal, whereby listener agent selects the perceived referent from a grid of 4 options. On each trial, listener and speaker receive negative or positive feedback on the guess. Each form-meaning pair is presented 3 times for each agent in each condition (speaking, listening), for a total of XX trials.

## Model specification
The model is designed to capture how iconicity can emerge in a dyadic communcication game through communicative pressure and incremental learning. The central assumption is that successful interaction favors form-meaning mappings that are iconic (perceptually transparent?), and that interlocutors adjust their behavior over repeated interactions to produce more iconic signals using both segmental (e.g., vowel quality, vowel duration) and suprasegmental acoustic properties (e.g., pitch, word length). 

Agents interact in a game in which meanings are conveyed using acoustic signals. Learning is gradual and experience-dependent: agents continuously update their internal representations of form-meaning associations based on interaction outcomes (success-failure). Learning is probabilistic, such that increases in learning do not guarantee perfect performance on every trial???

An additional assumption of the model is that the adaptive behavior of interlocutors happens through a learning algorithm whereby agents continuously learn to produce and perceive more iconic signals across the experiment. Learning is operationalized as gradual increase in associative strength, which probabilistically translate into correct recognition... This initial model is a simplification of the actual experiment outlined above, however, with the hypothesized psychological mechanisms in place. For instance, the initial learning stage of the model is lower than the threshold described above, and the perceptual acoustic space of agents is reduced to a two-dimensional space (see below).

Next, we describe the formalized parameters of the model, beginning with a description of agents and representations.

### Agents, signal and meaning spaces
  * there are two agents interacting in a language learning game, these agents behave rationally (?)
  * form-meaning pairs consists of a semantic referent and a signal; meanings are discrete categories (small vs. big), and signals are represented as points in a two-dimensional space
  
Meanings are discrete semantic categories corresponding to size objects (small vs big). These categories are assumed to be psychologically basic and perceptually salient, consistent with extensive evidence that magnitude distinctions are readily mapped onto acoustic dimensions.

Signals are represented as points in a continuous two-dimensional signal space. This space abstracts over multiple correlated acoustic dimensions (e.g., vowel quality, duration, pitch) that jointly contribute to perceived magnitude. The two dimensions do not correspond to specific acoustic parameters but capture the idea that signals vary continuously along perceptual dimensions relevant for iconic mapping.

Each meaning is associated with an iconic prototype in signal space. The small category is associated with a prototype at [0,0], and the big category with a prototype at [1,1]. These prototypes represent perceptually transparent signal–meaning correspondences rather than learned conventions.

---

While the target experiment consists of XX form-meaning pairs, the form-meaning pairs in the model presented here is reduced to a combination of two vectors: one with semantic category information---small, big---one with signal information, represented in a two-dimensional contained space where the prototypical signal is either [0,0] or [1,1]. The prototypical form-meaning pairs of the artificial language thus consist of a small referent paired with a [0,0] signal, and a large referent paired with a [1,1] signal. This two-dimensional signal space can be considered an abstraction over multiple correlated acoustic cues that jointly encode size-related meaning. This also implies that the task itself is reduced to a 2AFC task.

### Recognition and learning assumptions
  * learning occurs incrementally on every trial, and is modulated by success or failure


Agents maintain graded internal representations of form–meaning associations, formalized as guess strength. Guess strength reflects the agent’s subjective confidence or associative strength linking a signal to a meaning, rather than categorical accuracy.

On each trial, the listener attempts to recognize the referent intended by the speaker’s signal. Recognition is probabilistic: higher guess strength increases the likelihood of correct recognition, but does not guarantee it. Overt recognition decisions are categorical, reflecting the forced-choice nature of the task, while underlying representations remain continuous.

Learning occurs on every interaction. Both successful and unsuccessful trials provide information that updates the agent’s internal representations, though successful interactions contribute more strongly to learning. This assumption reflects error-driven and statistical learning mechanisms, whereby prediction errors and feedback incrementally strengthen associations over time.

---

Recognition and learning of the form-meaning pairs is formalized as guessing rate, which reflects the listener's subjective confidence in a form-meaning association. This increases through experience but does not guarantee correct behavior on every trial. 

At each trial, the listener guesses what referent the signal refers to. If the guess strength $> .5$, the guess is considered correct. The speaker then retrieves the signal for the referent and the signal remains the same. If the guess $< .5$, the guess is considered incorrect, there is no recall from memory and the speaker draws a new signal near [.5,.5] in the two-dimensional space. Although internal representations are continuous, responses are categorical in this 2AFC-task; a threshold on guess strength formalizes this decision boundary. Further, the .5 threshold reflects chance performance in a 2AFC-task.

The strength of guess rate is affected by the characteristics of the signal and the referent type (see below), but also of previous learning/listener memory. Previous learning equals the current guess rate of the agent. This implies that learning gets reinforced as higher previous guess rate generates higher chance of success on each trial, even before the speaker has produced a signal. Even unsuccessful trials provide information about the mapping space, leading to incremental learning, though with reduced learning rates relative to successful trials. This mirrors assumptions in implicit statistical learning, error-driven learning and gradient feedback sensitivity (where negative feedback is still informative).

### Iconicity bias
The emergence of iconicity in the model is driven by a learnability bias favoring signal–meaning mappings that are perceptually transparent. Signals that are closer to their iconic prototypes are easier to recognize and learn, resulting in higher initial guess strength and faster learning trajectories. Iconicity is thus not imposed directly, but arises from differential learnability under communicative pressure.

---

The emergence of iconicity is formalized as a learnability bias during guessing that favors signals that are more closely located to their prototypical target in the two-dimensional space. Some form-meaning pairs are thus easier to learn/guess depending on their spatial proximity to iconic prototypes. This assumes that signals closer to iconic prototypes are easier to perceive, encode, and retrieve, leading to higher initial guess strength and faster learning, making iconicity a processing advantage more so than a special symbolic mechanism???

### Noise and uncertainty
  * noise perturbs signal perception and production

Signal perception and production are subject to uncertainty. This uncertainty reflects perceptual noise, memory limitations, and articulatory variability. Under conditions of uncertainty or communicative failure, agents tend to produce less specified signals, which are biased toward the center of the signal space. This central tendency represents regression toward an unmarked or default signal when confidence in a form–meaning mapping is low.

---

Guessing nevertheless involves uncertainty which leads to imperfect learning. This is formalized as the input being perturbed by noise.  This noise could encompass both motor noise in production, external and internal noise in perception, as well as participants lapsing on certain trials (randomly responding without taking the signal into account). On trials where interaction fails, noise is formalized as signal drifting away from target, toward the middle of the two-dimensional space [.5,.5]; i.e., a regression toward an unmarked signal under uncertainty.

## Model implementation
Next, we describe how the model is implemented, step-wise going through the model.

Exact update equations for guess strength

Learning rates or increment sizes

Noise distributions (e.g., Gaussian, uniform)

Threshold values (e.g., > .5)

Number of trials, dyads, or simulation runs

Initialization values

Random seed handling

### Learning = the probability of correct guess
Learning of form-meaning pairs is implemented by calculating the probability of the correct guess at each trial in the interaction game. The probability of correct guess is a function of the previous guessing rate of agent (= the learned ability), the signal fit (= signal ease) and iconicity bias (= iconicity boost). 

The initial learning state of the learner (after training) is summarized to an averaged probability of ~.33. The probability is established by drawing n random values from a beta distribution with a mu = .333 and sigma = .032.

The signal fit is a distance-based measure of Euclidean distance that expresses target distance to attractor, i.e., the iconic prototype. If a signal is closer to its prototype, it is easier to guess (for small referents, closer to [0,0], for large referents, closer to [1,1]). The distance measure drops off sharply (exponentially) when moving away from targets.
The signal fit is multiplied with an iconicity boost that represents the strength of iconicity affecting guessing (a constant that can be changed by the researcher to account for the strength of iconicity). This means that iconicity bias is implemented as reinforcing the signal fit to iconic target---some signals are easier to guess depending on their spatial proximity to iconic prototypes. 

The probability is bounded at $pmax = .95$ to account for $.05$ noise/lapse. At each trial, the posterior probability is implemented as drawing from a random beta distribution. After each trial, a learning rate of $p = .005$ is added to the listener guess, regardless of guess being correct or incorrect. If the learner guess is correct, the same rate is added to the listener guess again, thus increasing learning when correct. The signal, however, remains intact.<!--psych mech of learn rate: plasticity/memory strenght?-->

Cultural selection bias: implemented as a reinforcement rate on successful trials. The learned/stored signal is updated toward the speaker's produced signal by a fixed proportion. This biases learning toward forms that have been successful in communication, strengthening emergent conventions over time (by reinforcing forms that are already successful in the community and consolidating new forms). It affects memory update on successful trials, moving the stored signal toward the produced signal on success, affecting the shared memory of signals.<!--Is this a strategy of making the signal more iconic to make it more learnable in the communicative game? psych mech: social alignment?-->

### Signal update
At each trial, the signal updates depending on the speaker's current guess rate, prior to guessing. This production bias toward prototype in production updates the signal by moving it a small fixed step toward the iconic prototype for that referent type. This adds a directional pressure pushing small signals to move towards [0,0] or large signals to move towards [1,1], i.e., push of signals towards the iconic target (directional evolution). When speakers' guessing rates are higher, the bias is smaller and a signal closer to stored value is produced; when speakers' guessing rates are lower, bias is larger and speakers produce a more prototypical signal. The bias is applied prior to calculating the posterior probability, hence, it only affects the signal being produced not the stored representation.<!--psychological mechanism: motor simplification, audience design?--> This is implemented as sampling from beta?? distribution with a mean shifted toward [0,0] or [1,1]: this will affect signal generation becoming more iconic over time.

In addition to this deterministic shift, two-dimensional Gaussian noise is added, sampled from a distribution with $\mu = 0$, $\sigma = .05$, and bounded to [0,1]. Thus, signal drift on failure is directional toward the prototype (iconicity) but includes random variation from noise. The failure drift could be interpreted as an articulatory bias, i.e., an intentional iconicity parameter on behalf of the speaker when noticing that communication fails.<!--Given that there are no minimal pairs of evident contrasts in the experiment, it might be that this parameter generates iconicity even if the real design could not produce this mechanism?; psych mech: error-based learning/repair?-->

## Results of simulation
Here, we present the results of a simulation of XX number of dyads, XX number of trials. Random seed handling.


## Concerns
  * learning update: the guess probabilities improve over time while signal drift decreases (no drift on correct trials), this means that the signal tends to freeze near its initial state [.5,.5].
  * the iconicity bias: iconicity is implemented through the ease function which improves listeners' probability of correct guesses, but it does not affect the likelihood of the speaker producing good-guess signals.
  * add cultural selection bias where successful signals get reinforced: this affects memory update

Is there a memory bias in the model? There seems to be an implicit benefit of learning iconic forms quicker, but without making iconicity emerge? Success in the model does not directly improve memory of the signal form. It only improves the guess probability but not the stored signal. So, iconic forms are easier to understand but do not improve memory of the signal form. However, there is direct memory bias on unsuccessful trials, the update_signal_on_failure function moves memory toward prototype/iconic patterns. If communication fails, the stored signal shifts toward the iconic prototype---direct iconicity bias. There is also indirect memory bias using prod_bias because produced signals trend toward prototypes for speakers with lower guess rate, and even for speakers with high guess rates when noise pushes them. Those produced signals then get reinforced on success by reinforce_signal. So the memory drift and reinforcement loop leads to: production drifting toward iconic prototypes, those forms are easier to interpret (signal_ease), they yield more successes, success to reinforcement to memory updates toward what was produced.

Setting production bias = 0, reinforcement rate = 0, iconicity boost = 0 has no effect on the emergence of iconicity. Setting failure_step > 0, however, introduces iconicity. This seems to suggest that the strongest effect on the emergence of iconicity is in whether the signal drifts toward the prototype on incorrect trials.

```{r}
# show heatmap of ease relative to signal space
grid <- expand.grid(
  x = seq(0, 1, length.out = 50),
  y = seq(0, 1, length.out = 50),
  type = c("small", "large"))

# make heat map
grid <- grid %>%
  rowwise() %>%
  mutate(ease = signal_ease(c(x, y), type)) %>%
  ungroup()

ggplot(grid,
       aes(x = x, y = y, fill = ease)) +
  geom_tile() +
  facet_wrap(. ~type) +
  scale_fill_viridis_c()

```


```{r parameters}
set.seed(2139752)

# Generate empty dataframe for simulation
d.empty <- data.frame(
  sim      = integer(),
  round    = integer(),
  referent = integer(),
  speaker  = character(),
  listener = character(),
  type     = character(),
  p_correct = numeric(),
  success   = integer(),
  stored_x  = numeric(),
  stored_y  = numeric(),
  produced_x = numeric(),
  produced_y = numeric(),
  stringsAsFactors = FALSE)

# If in logodds space, the learnrate should change from .005
learn_rate_prob <- 0.005
# Approx starting point for agents
p0 <- 0.3
learn_rate_logit <- qlogis(p0 + learn_rate_prob) - qlogis(p0)

# Generate one data frame with different simulations where params are switched on/off
d.simulation <- rbind(
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, learn_rate = learn_rate_logit) %>%
    mutate(parameter_space = "allParams"),
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, learn_rate = learn_rate_logit, prod_bias = 0) %>%
    mutate(parameter_space = "noProdBias"),
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, learn_rate = learn_rate_logit, failure_step = 0) %>%
    mutate(parameter_space = "noFailureStep"),
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, learn_rate = learn_rate_logit, reinforcement_rate = 0) %>%
    mutate(parameter_space = "noReinforcement"),
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, learn_rate = 0) %>%
    mutate(parameter_space = "noLearning"),
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, learn_rate = learn_rate_logit, iconicity_boost = 0) %>%
    mutate(parameter_space = "noIconicityBoost"),
  # Also add one simulation that resembles the old model
  d.empty %>% run_interaction_sim(n_sim = 10, n_rounds = 1000, learn_rate = learn_rate_logit, prod_bias = 0, failure_step = 0, reinforcement_rate = 0) %>%
    mutate(parameter_space = "onlyLearning"))

# Diagnostics
# d.onlyDrift <- d.empty %>%
#   run_interaction_sim(n_sim = 10, n_rounds = 1000, learn_rate = 0, prod_bias = 0, reinforcement_rate = 0, iconicity_boost = 0) %>%
#   mutate(parameter_space = "onlyDrift")

```

```{r iconicity}
# you can compute iconicity either from stored signals or produced tokens
d.iconicity <- d.simulation %>%
  group_by(parameter_space) %>%
  mutate(bins = cut(round, breaks = 20, labels = FALSE)) %>%
  group_by(parameter_space, bins, type, sim) %>%
  summarise(
    x_learned = mean(stored_x),
    y_learned = mean(stored_y),
    x_produced = mean(produced_x),
    y_produced = mean(produced_y),
    .groups = "drop") %>%
  pivot_longer(
    cols = starts_with(c("x", "y")),
    names_to = c("signal", "process"),
    values_to = "value",
    names_sep = "_") %>%
  pivot_wider(
    names_from = signal,
    values_from = value) %>%
  mutate(
    target_x = ifelse(type == "small", 0, 1),
    target_y = ifelse(type == "small", 0, 1),
    dist = sqrt((x - target_x)^2 + (y - target_y)^2),
    iconicity = exp(-2 * dist)) %>%
  group_by(parameter_space, bins, sim, process) %>%
    summarise(iconicity = mean(iconicity), .groups = "drop")

```

```{r accuracy, fig.width=base.width*6, fig.height=base.height*3, fig.align='center', out.width='90%'}
# Plot success rate over time across simulation within parameter space
d.simulation %>%
  #group_by(parameter_space, round) %>%
  #summarise(success = mean(success)) %>%
  ggplot(
    aes(x = round, y = success)) +
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"), 
              se = TRUE) +
  labs(title = "Success rate over time (across simulations)", 
       y = "Success", 
       x = "Round") +
  facet_grid( ~ parameter_space) +
  scale_y_continuous(limits = c(0,1)) + 
  theme_minimal()
```

```{r iconicity-plot, fig.width=base.width*6, fig.height=base.height*3, fig.align='center', out.width='90%'}
#iconicity trajectories
p.learned.iconicity <- d.iconicity %>%
  filter(process == "learned") %>%
  group_by(parameter_space, bins, sim) %>% 
  summarise(iconicity = mean(iconicity)) %>%
  ggplot(
    aes(x = bins, y = iconicity, group = sim)) +
  geom_path(size = 0.9, alpha = 0.2) +
  geom_path(data = . %>% 
              group_by(parameter_space, bins) %>%
              summarise(iconicity = mean(iconicity)), 
            aes(group = 1), size = 1.5) +
  geom_hline(yintercept = exp(-2 * sqrt((0.5)^2 + (0.5)^2)), lty = "dashed") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
  labs(title = "Iconicity over time (learned tokens)",
       y = "Iconicity (high = iconic)", x = "Bins (rounds in 100s)") +
  theme_minimal() +
  facet_grid(~parameter_space)

# p.produced.iconicity <- p.learned.iconicity %+%
#   (d.iconicity %>%
#   filter(process == "produced") %>%
#   group_by(parameter_space, bins, sim) %>% 
#   summarise(iconicity = mean(iconicity))) +
#   labs(title = "Iconicity over time (produced signals)") +
#   facet_grid(~parameter_space)

p.learned.iconicity

d.simulation %>%
  mutate(
    dx = produced_x - stored_x,
    dy = produced_y - stored_y
  ) %>%
  group_by(parameter_space, success) %>%
  summarise(
    mean_step = mean(sqrt(dx^2 + dy^2)),
    .groups = "drop"
  )

```

### Accuracy-iconicity dissociation
Learning (success over time) only fails when learn_rate is switched off, suggesting that guessing accuracy depends entirely on learn_rate (expected give that it is the only mechanism that updates guessing probabilities). Iconicity fails in two cases: when there is only learning and iconicity boost, and when there is no signal drift toward prototype on failed trials. This seems to suggest that iconicity is not a by-product of communicative success, and that it requires specific signal-space dynamics.

Is there learning 'for free' even without iconicity bias? Check whether accuracy mediates iconicity (post-hoc diagnostics). A) where iconicity is impossible (so when only learn_rate and noise is in the model), check whether separation can be predicted by accuracy + learn_rate + noise. B) where iconicity is possible, check whether separation can be predicted by iconicity + accuracy + failure_step.

TODO: check iconicity separately for success and fail trials (will no failure model show that iconicity only accumulates after success?, will failure models show that iconicity jumps after failures?)

```{r fig.width=base.width*6, fig.height=base.height*3, fig.align='center', out.width='90%'}
# Signal space use
p.learned.signals <- d.simulation %>%
  ggplot(
    aes(stored_x, 
        stored_y, 
        color = type)) +
  geom_point(alpha = 0.2) +
  labs(title = "Evolution of learned signal space", x = "x", y = "y") +
  facet_grid(parameter_space ~ sim) +
  scale_x_continuous(limits = c(0,1)) + 
  scale_y_continuous(limits = c(0,1)) + 
  scale_color_viridis_d(begin = 0.1, end = 0.9) +
  theme_minimal()

p.produced.signals <- p.learned.signals %+%
  aes(produced_x,
      produced_y) +
  labs(title = "Evolution of produced signal space", x = "x", y = "y")

p.learned.signals + p.produced.signals +
  plot_layout(guides = "collect")

```

Figure XX seems to suggest that categories separate even for model runs where iconicity does not emerge. Why might this be the case? Could it be that separation is emerging from Bayesian updating + task structure? informative feedback structure (posteriors concentrate around category-consistent regions, inducing implicit clustering?)? Shared target structure (distinct targets, repeated sampling?)? Asymmetric error correction (incorrect guesses trigger larger updates than correct guesses --> category divergence)?
  * replace informative feedback with random correctness and check whether separation disappears (feedback structure is causal).
  * set learn_rate = 0 but allow guesses to improve (inference alone is sufficient).

## Grid over the parameter space
In order to investigate what parameter settings generate the largest overall iconicity across trials and simulations, we generated a grid search over four model parameters: production bias toward prototypes, strength of iconicity in comprehension (iconicity boost), strength of cultural selection (reinforcement rate), rate of learning in log-odds space (learning rate) and signal drift toward iconic prototypes on incorrect trials. The grid search runs the interaction model for each combination, and computes how iconic the signals become by the end of learning.

Figure XX suggest that the more I learn for free (learn_rate), the less iconic I become. This would suggest that iconicity might not continue to grow in a language, at least not as a learning mechanism (only as a strategy to make the language more learnable for the interlocutor).

We also note that under some condition, iconicity still emerges when signal drift on failed signals is switched off (panel XX, Fig XX). This suggests that failure-based drift is not strictly necessary but is normally the dominant pathway (error-driven vs. success-driven iconicity). Generating a positive feedback-loop of biased production--higher success--reinforcement--more biased stored signals--higher success..

To explore: a memory bias toward more iconic forms? Making the signals more iconic does not seem to make iconicity emerge, merely increase learning rate. Currently, the model does not implement a mechanism whereby the memory of the signal form is reinforced as a function of its iconicity directly. Iconicity improves the guess probability but not the stored representation. That is, iconic forms are easier to understand but they do not improve memory of the signal form. However, there is direct memory bias on failure--the update_signal_on_failure function moves memory representation toward prototype on incorrect trials. Indirectly, there is iconicity bias on produced signals--the produced signals get reinforced on success, leading to production drifting toward iconic prototypes. Those forms are easier to interpret (signal_ease), yielding more successes and memory updates towards what was produced (so indirect bias via production bias).

```{r}
if(RESET_MODELS || !file.exists("../models/grid-search.rds")) {
  
  # Generate a parameter grid to explore
  d.param_grid <- expand.grid(
    prod_bias = seq(0, 0.3, length.out = 4),
    iconicity_boost = seq(0, 0.4, length.out = 4),
    reinforcement_rate = seq(0, 0.1, length.out = 4),
    failure_step = seq(0, 0.06, length.out = 4),
    learn_rate = seq(0, 0.045, length.out = 4))
  
  # Prepare results container
  d.grid_results <- d.param_grid %>%
    mutate(iconicity = NA_real_,
           history = vector("list", n()))
  
  for (i in seq_len(nrow(d.param_grid))) {
    
    params <- d.param_grid[i, ]
    
    # empty structure for simulation log
    empty_df <- data.frame(
      sim = integer(), round = integer(), referent = integer(),
      speaker = character(), listener = character(), type = character(),
      p_correct = numeric(), success = integer(),
      stored_x = numeric(), stored_y = numeric(),
      produced_x = numeric(), produced_y = numeric(),
      stringsAsFactors = FALSE
    )
    
    # run simulation
    history <- run_interaction_sim(
      data = empty_df,
      n_sim = 10,
      n_referents = 6,
      n_rounds = 1000,
      drift_sd = 0.05,
      learn_rate = params$learn_rate,
      iconicity_boost = params$iconicity_boost,
      prod_bias = params$prod_bias,
      reinforcement_rate = params$reinforcement_rate,
      failure_step = params$failure_step,
      lapse = 0.05)
    
    # store full dataframe
    d.grid_results$history[[i]] <- history
    
    # compute iconicity score
    d.grid_results$iconicity[i] <- compute_iconicity(history)
    
    message("Completed parameter set ", i, " of ", nrow(d.param_grid))
  }
  
  
  saveRDS(d.grid_results, file = "../models/grid-search.rds", compress = T)
  
} else { d.grid_results <- readRDS("../models/grid-search.rds") }

d.full.history <- bind_rows(d.grid_results$history, .id = "param_set")

```
```{r fig.width=base.width*4, fig.height=base.height*3.5, fig.align='center', out.width='90%'}
# plot for different param settings
p.noReinforce <- d.grid_results %>%
  mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
  filter(reinforcement_rate == 0) %>%
  #filter(failure_step == 0) %>%
  rename(`signal drift` = failure_step, `production bias` = prod_bias) %>%
  ggplot(
    aes(
      x = learn_rate, 
      y = iconicity_boost, 
      fill = iconicity,
      color = ifelse(iconicity != max(iconicity), "red", "black"))) +
  geom_tile() +
  geom_tile(
    data = . %>%
      filter(iconicity >= 0.24 & iconicity <= 0.26),
    color = "yellow",
    size = .3,
    fill = NA) +
  geom_point(
    data = . %>%
      filter(iconicity == max(iconicity)),
    shape = 4) +
  scale_color_manual(values = c("red", "black")) +
  scale_fill_viridis_c() +
  facet_grid(`production bias` ~ `signal drift`, labeller = label_both) +
  theme_minimal() +
  guides(color = "none") +
  labs(
    title = "Iconicity across the parameter space for when there is no reinforcement of successful signals",
    x = "Learning rate",
    y = "Iconicity boost",
    fill = "Iconicity")

p.noProdBias <- p.noReinforce %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(prod_bias == 0) %>%
     rename(`signal drift` = failure_step, reinforcement = reinforcement_rate)) +
  facet_grid(reinforcement ~ `signal drift`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space for when there is no production bias")

p.noDrift <- p.noReinforce %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(failure_step == 0) %>%
     rename(reinforcement = reinforcement_rate, `production bias` = prod_bias)) +
  facet_grid(reinforcement ~ `production bias`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space for when there is no signal drift")

p.NoLearning <- p.noReinforce %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(learn_rate == 0) %>%
     rename(reinforcement = reinforcement_rate, `signal drift` = failure_step)) +
  aes(x = prod_bias) +
  facet_grid(reinforcement ~ `signal drift`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space for when there is no learning") +
  scale_x_continuous("Production bias (toward target)")

p.NoBoost <- p.noReinforce %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(iconicity_boost == 0) %>%
     rename(`reinforcement` = reinforcement_rate, `signal drift` = failure_step)) +
  aes(y = prod_bias) +
  facet_grid(reinforcement ~ `signal drift`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space for when there is no iconicity boost") +
  scale_y_continuous("Production bias (toward target)")

(p.noReinforce + p.noProdBias) /
  (p.noDrift + p.NoLearning + p.NoBoost) +
  plot_layout(guides = "collect")

```


```{r fig.width=base.width*4, fig.height=base.height*3.5, fig.align='center', out.width='90%'}
# # Check params relative weight
# m.params <- lm(iconicity ~ prod_bias + iconicity_boost + learn_rate + reinforcement_rate + failure_step,
#    data = d.grid_results) %>%
#   tidy()

# plot for different param settings
p.grid.noBoost <- d.grid_results %>%
  mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
  filter(iconicity_boost == 0) %>%
  rename("reinforcement" = reinforcement_rate, "production bias" = prod_bias) %>%
  ggplot(
    aes(
      x = failure_step, 
      y = learn_rate, 
      fill = iconicity,
      color = ifelse(iconicity != max(iconicity), "red", "black"))) +
  geom_tile() +
  geom_point(
    data = . %>%
      filter(iconicity == max(iconicity)),
    shape = 4) +
  scale_color_manual(values = c("red", "black")) +
  scale_fill_viridis_c() +
  facet_grid(reinforcement ~ `production bias`, labeller = label_both) +
  theme_minimal() +
  guides(color = "none") +
  labs(
    title = "Iconicity across the parameter space for when there is no iconicity boost",
    x = "Signal drift on failed trials (toward prototype)",
    y = "Learning rate",
    fill = "Iconicity")

p.grid.noProdBias <- p.grid.noBoost %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(prod_bias == 0) %>%
     rename(reinforcement = reinforcement_rate, `Iconicity boost` = iconicity_boost)) +
  facet_grid(reinforcement ~ `Iconicity boost`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space for when there is no production bias")

p.grid.noReinforcement <- p.grid.noBoost %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(reinforcement_rate == 0) %>%
     rename(`Production bias` = prod_bias, `Iconicity boost` = iconicity_boost)) +
  facet_grid(`Production bias` ~ `Iconicity boost`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space for when there is no signal reinforcement")

p.grid.NoLearning <- p.grid.noBoost %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(learn_rate == 0) %>%
     rename(reinforcement = reinforcement_rate,
            `Iconicity boost` = iconicity_boost)) +
  aes(y = prod_bias) +
  facet_grid(reinforcement ~ `Iconicity boost`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space for when there is no learning") +
  scale_y_continuous("Production bias (toward target)")

p.grid.NoDrift <- p.grid.noBoost %+%
  (d.grid_results %>%
     mutate(across(where(is.numeric), ~ round(.x, digits = 3))) %>%
     filter(failure_step == 0) %>%
     rename(`reinforcement` = reinforcement_rate, `Iconicity boost` = iconicity_boost)) +
  aes(x = prod_bias) +
  facet_grid(reinforcement ~ `Iconicity boost`, labeller = label_both) +
  labs(title = "Iconicity across the parameter space for when there is no signal drift") +
  scale_x_continuous("Production bias (toward target")

(p.grid.NoDrift + p.grid.NoLearning) /
  (p.grid.noProdBias + p.grid.noReinforcement + p.grid.noBoost) +
  plot_layout(guides = "collect")

```

Also, plot grid results where all other params are at empirically reasonable values, and only vary one parameter at the time. Check whether iconicity even appears without it, and whether accuracy depend on it. This will allow me to distinguish between necessary and sufficient params.

One way to evaluate this is visualized in Figure XX below. Here, each panel shows the data from the grid search down-sampled to when the labelled parameter varies and the other parameters are set to a 'plausible' value: for production bias = 0.2, for iconicity boost = 0.267, for reinforcement rate = 0.067, for failure drift = 0.04, and for learning rate = 0.03. So, given a baseline configuration where other parameters are reasonable, how does iconicity change as I increase this parameter? If iconicity fails to emerge when the parameter value is at 0, no matter what the other parameters are doing, it is likely a necessary parameter. If iconicity emerges when the parameter > 0 even when all other parameters are set to 0, then it is likely a sufficient parameter.

Figure XX suggests that only signal drift on failed trials is likely to be a necessary parameter. It also seems to be a sufficient parameter as when setting all other parameters = 0, iconicity still emerges. None of the other parameters seem to be sufficient, though.

Without signal drift, reinforcement only strengthens already successful forms and production bias is weak when speaker guess increases. In addition, early random successes can lock in arbitrary, non-iconic equilibria, meaning that whatever random form works first gets reinforced even if it is not iconic.

```{r}
# #Generate temp dataframe
d.grid_results.ic <- d.grid_results %>%
  unnest(history)

# Compute accuracy for each simulation, parameter set, and optionally per round
d.grid_results.ic.summary <- d.grid_results.ic %>%
  group_by(prod_bias, iconicity_boost, reinforcement_rate, failure_step, learn_rate) %>%
  summarise(
    accuracy = mean(success),   # proportion correct across all sims and rounds
    iconicity = unique(iconicity), # if iconicity already stored per param set
    .groups = "drop")

# Generate one dataframe where only one of the params are evaluated at a time
d.iconicity.grid <- d.grid_results.ic.summary %>%
  filter(between(iconicity_boost, 0.2, 0.3) & between(reinforcement_rate, 0.06, 0.09) & failure_step == 0.04 & learn_rate == 0.03) %>%
  mutate(parameter_space = "productionBias",
         value = prod_bias) %>%
  select(parameter_space, value, iconicity, accuracy) %>%
  rbind(d.grid_results.ic.summary %>%
          filter(between(prod_bias, 0.1, 0.3) & between(reinforcement_rate, 0.06, 0.09) & failure_step == 0.04 & learn_rate == 0.03) %>%
          mutate(parameter_space = "iconicityBoost",
                 value = iconicity_boost) %>%
          select(parameter_space, value, iconicity, accuracy),
        d.grid_results.ic.summary %>%
          filter(between(iconicity_boost, 0.2, 0.3) & between(prod_bias, 0.1, 0.3) & failure_step == 0.04 & learn_rate == 0.03) %>%
          mutate(parameter_space = "reinforcementRate",
                 value = reinforcement_rate) %>%
          select(parameter_space, value, iconicity, accuracy),
        d.grid_results.ic.summary %>%
          filter(between(iconicity_boost, 0.2, 0.3) & between(reinforcement_rate, 0.06, 0.09) & between(prod_bias, 0.1, 0.3) & learn_rate == 0.03) %>%
          mutate(parameter_space = "failureStep",
                 value = failure_step) %>%
          select(parameter_space, value, iconicity, accuracy),
        d.grid_results.ic.summary %>%
          filter(between(iconicity_boost, 0.2, 0.3) & between(reinforcement_rate, 0.06, 0.09) & failure_step == 0.04 & between(prod_bias, 0.1, 0.3)) %>%
          mutate(parameter_space = "learnRate",
                 value = learn_rate) %>%
          select(parameter_space, value, iconicity, accuracy))
# 
# d.iconicity.grid.long <- d.iconicity.grid %>%
#   pivot_longer(
#     cols = c(prod_bias, iconicity_boost, reinforcement_rate, failure_step, learn_rate),
#     names_to = "parameter",
#     values_to = "value")

d.iconicity.grid %>%
  ggplot(
    aes(
      x = value, 
      y = iconicity)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se,
               geom = "ribbon",
               alpha = 0.2) +
  facet_wrap(~ parameter_space, scales = "free_x") +
  labs(
    x = "Parameter value",
    y = "Mean iconicity",
    title = "Which parameters are necessary or sufficient for iconicity?") +
  theme_minimal()

d.iconicity.grid %>%
  ggplot(
    aes(
      x = value, 
      y = accuracy)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.2) +
  facet_wrap(~ parameter_space, scales = "free_x") +
  labs(x = "Parameter value", y = "Mean accuracy") +
  theme_minimal()



#Necessary param
d.grid_results %>%
  filter(failure_step == 0) %>%
  summarise(mean_iconicity = mean(iconicity))

#sufficent param
d.grid_results %>%
  filter(
    failure_step == 0,
    prod_bias == 0,
    reinforcement_rate == 0,
    iconicity_boost == 0,
    learn_rate >= 0)
```
