---
title: "Supplementary Information"
format: html
editor: source
---

# Supplementary Information for Persson & Roettger, 2026

This is the Supplementary Information (SI) for the XX paper. The SI contains models and analyses...

## Calibration of parameter values

...calibration based on theoretical plausibility/deterministic math...

```{r load-libraries, echo=FALSE}
library(dplyr)
library(purrr)
library(tibble)
library(ggplot2)
library(scales)
library(tidyr)
```

### Attractor dimensions and mechanics: circle radius, sd of produced signal, sd of signal inside attractor

The attractor dynamics is defined by three parameters: the circle radius, the standard deviation of the two-dimensional signal that is being produced, and the standard deviation of a signal once inside an attractor.

```{r functions, echo=FALSE}
# core helpers from main simulation function
clamp01 <- function(x) pmax(0, pmin(1, x))
clamp02 <- function(x) pmax(0, pmin(0.95, x))

produce_signal <- function(stored_signal, speaker_guess, drift_sd, k_production,
                           attractor_centers, circle_radius, center_sd,
                           k_attractor_production) {
  baseline_sd <- drift_sd * (1 + k_production * (0.5 - speaker_guess))
  distances <- sapply(attractor_centers, function(center) sqrt(sum((stored_signal - center)^2)))
  dist_to_nearest <- min(distances)
  nearest_id <- which.min(distances)
  inside_attractor <- distances < circle_radius
  is_inside <- any(inside_attractor)
  
  max_possible_baseline_sd <- drift_sd * (1 + k_production * 0.5)
  f_dist_max <- max_possible_baseline_sd - center_sd
  attractor_reduction <- 0
  
  if (is_inside) {
    closest_attractor_dist <- min(distances[inside_attractor])
    rel_dist <- closest_attractor_dist / circle_radius
    magnitude <- exp(-k_attractor_production * rel_dist)
    max_mag <- exp(-k_attractor_production * 0)
    min_mag <- exp(-k_attractor_production * 1)
    exp_scale <- (magnitude - min_mag) / (max_mag - min_mag)
    attractor_reduction <- f_dist_max * exp_scale
  }
  
  final_sd <- max(baseline_sd - attractor_reduction, 0.001)
  signal <- rnorm(length(stored_signal), mean = stored_signal, sd = final_sd)
  
  list(signal = clamp01(signal),
       inside_attractor = is_inside,       # renamed to match main script
       dist_to_nearest = dist_to_nearest,  # added, was missing
       attractor_id = if (is_inside) nearest_id else NA_integer_)
}

```

The circle radius of the attractors were set to 0.3, given the boundaries of the space (0-1 in both dimensions).

The sd of the produced signal (drift_sd) was calibrated to 0.2, which corresponds to a \~43% chance of the signal wandering into any attractor in a single production step at the initial stages of the interaction—when the signal is neutral \[0.5, 0.5\], and the signal sender's associative strength (speaker_guess) is 0.3. The calibration was done using a function that generates single-step signal productions from the signal origin \[0.5, 0.5\] and calculates the probability of a signal landing within the defined boundary radius of any attractor basin (0.3).

```{r}
test_single_step_capture <- function(attractor_centers = list(c(0.15, 0.85), c(0.85, 0.15), 
                                                             c(0.15, 0.15), c(0.85, 0.85)),
                                      circle_radius = 0.3,
                                      drift_sd = 0.2,
                                      k_production = 1.5,
                                      guess = 0.5,
                                      start = c(0.5, 0.5),
                                      n_sims = 100000,
                                      movement = "signal") {
  # Effective standard deviation
  baseline_sd <- drift_sd * (1 + k_production * (0.5 - guess))
  # generate 2D signal draws
  sig_x <- pmax(0, pmin(1, rnorm(n_sims, mean = start[1], sd = baseline_sd)))
  sig_y <- pmax(0, pmin(1, rnorm(n_sims, mean = start[2], sd = baseline_sd)))
  # adjust target point depending on whether testing raw signal or half-step movement
  if (movement == "half_step") {
    target_x <- (sig_x + start[1]) / 2
    target_y <- (sig_y + start[2]) / 2
  } else {
    target_x <- sig_x
    target_y <- sig_y
  }
  # check capture across all attractors
  captured <- rep(FALSE, n_sims)
  for (c in attractor_centers) {
    dist_sq <- (target_x - c[1])^2 + (target_y - c[2])^2
    captured <- captured | (dist_sq < circle_radius^2)
  }
  
  mean(captured)
}

attractors <- list(c(0.15, 0.85), c(0.85, 0.15), c(0.15, 0.15), c(0.85, 0.85))
# Test raw signal capture
p_signal <- test_single_step_capture(
  attractor_centers = attractors,
  circle_radius = 0.3,
  drift_sd = 0.2,
  movement = "signal")
cat("Single-step signal capture probability:", round(p_signal, 3), "\n")

```

The sd of a signal that has fallen into an attractor (center_sd) was calibrated to 0.12, which corresponds to a \~2% chance of escaping from the attractor in a single production step when the signal is at the dead center and the circle_radius is 0.3. The dead center of an attractor is therefore extremely sticky; however, the probability of hitting the dead center from the start of the interaction is low (\~5% chance of landing within a tight radius of $r \le 0.10$). Signals close to the boundary of an attractor have a sd close to baseline noise (drift_sd), making them likely to both drift deeper into the attractor or escape it entirely. We ran a simulation exploring how the probability of escaping an attractor is affected by signal location and associative strength. The simulation suggests that escape probability from the dead center is barely affected by associative strength, whereas its effect is substantially larger for off-center signals--the escape probability for signals located at a quarter-radius drops from 0.29 to 0.055 as associative strength increases (from 0 to 1). Signals at the boundary retain a high escape probability (\>0.56 across associative strengths).

```{r}
test_single_step_escape <- function(attractor_center = c(0.15, 0.15),
                                    circle_radius = 0.3,
                                    center_sd = 0.12,
                                    drift_sd = 0.2,
                                    k_production = 1.5,
                                    guess = 0.2,
                                    n_sims = 100000) {
  final_sd <- max(center_sd - drift_sd * k_production * guess, 0.001)
  
  sig_x <- pmax(0, pmin(1, rnorm(n_sims, mean = attractor_center[1], sd = final_sd)))
  sig_y <- pmax(0, pmin(1, rnorm(n_sims, mean = attractor_center[2], sd = final_sd)))
  
  dist_sq <- (sig_x - attractor_center[1])^2 + (sig_y - attractor_center[2])^2
  mean(dist_sq > circle_radius^2)
}

test_single_step_escape(center_sd = 0.12, guess = 0)  # should closely match escape_probability_exact(0.12) above

# Code kept for reproducibility
# # Test the probability of hitting the dead center within one production step at the start of the interaction
# test_center_capture_probability <- function(n_reps = 100000, speaker_guess = 0.3) {
#   stored_signal <- c(0.5, 0.5)
#   circle_radius <- 0.30
#   
#   # Full list of 4 attractor centers defined in run_interaction_sim
#   attractor_centers <- list(
#     c(0.15, 0.85),
#     c(0.85, 0.15),
#     c(0.15, 0.15),
#     c(0.85, 0.85))
#   
#   # Run Monte Carlo production draws
#   signals <- replicate(n_reps, {
#     produce_signal(
#       stored_signal = stored_signal,
#       speaker_guess = speaker_guess,
#       drift_sd = 0.2,
#       k_production = 1.5,
#       attractor_centers = attractor_centers,
#       circle_radius = circle_radius,
#       center_sd = 0.12,
#       k_attractor_production = 2.5
#     )$signal
#   })
#   
#   # Compute minimum distance to any attractor center for each trial
#   min_dists <- apply(signals, 2, function(sig) {
#     min(sapply(attractor_centers, function(ctr) sqrt(sum((sig - ctr)^2))))
#   })
#   
#   tibble(
#     stored_signal = "(0.5, 0.5)",
#     speaker_guess = speaker_guess,
#     effective_sd = 0.2 * (1 + 1.5 * (0.5 - speaker_guess)), # = 0.247
#     p_captured_any_basin = mean(min_dists < 0.30),
#     p_near_center_r010    = mean(min_dists < 0.10),
#     p_near_center_r005    = mean(min_dists < 0.05))
# }
# 
# set.seed(42)
# d.center_hit <- test_center_capture_probability(speaker_guess = 0.3)
# print(d.center_hit)

# calibrate_attractor_escape <- function(n_reps = 100000) {
#   attractor_center <- c(0.15, 0.15)
#   attractor_centers <- list(attractor_center, c(0.85, 0.85))
#   circle_radius <- 0.3
#   
#   test_grid <- expand.grid(
#     rel_distance = c(0.0, 0.25, 0.50, 0.75, 1), # 0 = center, 1 = boundary
#     speaker_guess = c(0.0, 0.25, 0.5, 0.75, 1)
#   )
#   
#   results <- test_grid %>%
#     mutate(
#       empirical_p_escape = map2_dbl(rel_distance, speaker_guess, function(rel_dist, guess) {
#         # Explicit 2D vector for stored signal position
#         stored_signal <- c(attractor_center[1] + (rel_dist * circle_radius), attractor_center[2])
#         
#         # Replicate 2D draws (returns a 2 x n_reps matrix)
#         sig_matrix <- replicate(n_reps, {
#           produce_signal(
#             stored_signal = stored_signal,
#             speaker_guess = guess,
#             drift_sd = 0.19,
#             k_production = 1.5,
#             attractor_centers = attractor_centers,
#             circle_radius = circle_radius,
#             center_sd = 0.12,
#             k_attractor_production = 2.5
#           )$signal
#         })
#         
#         # Distance of each produced signal from the attractor center
#         dists <- sqrt((sig_matrix[1, ] - attractor_center[1])^2 + 
#                       (sig_matrix[2, ] - attractor_center[2])^2)
#         
#         # Proportion escaping boundary
#         mean(dists >= circle_radius)
#       })
#     )
#   
#   return(results)
# }
# 
# # Run calibration
# set.seed(42)
# d.calibration <- calibrate_attractor_escape()

```

### Learning strength

After signal recognition, signal sender and signal receiver update their representations as dependent on interaction outcome---success or failure. The amount of learning is controlled by the parameters learning strength and success/failure scale. The learning strength is calibrated to 0.015 with a success scale of 7.5 and a failure scale of 1. This corresponds to a boost of \~0.02 for a successful trial and \~0.003 for a failed trial from the start of the interaction, and reaching an overall associative strength of \~0.96 after 50 trials of all-successes, and of \~0.48 after 50 trials of all-failures.

```{r}
# check the two idealized extreme trajectories (pure-success-only, pure-failure-only) for a given learning_strength/success_scale/failure_scale.
report_pure_paths <- function(learning_strength, success_scale = 7.5, failure_scale = 1,
                              n_updates = 50, start_p = 0.3) {
  logit_success <- qlogis(clamp02(start_p))
  logit_failure <- qlogis(clamp02(start_p))
  for (i in seq_len(n_updates)) {
    logit_success <- qlogis(clamp02(plogis(logit_success))) + (learning_strength * success_scale)
    logit_failure <- qlogis(clamp02(plogis(logit_failure))) + (learning_strength * failure_scale)
  }
  tibble(
    n_updates = n_updates,
    start_p = start_p,
    pure_success_p = plogis(logit_success),
    pure_failure_p = plogis(logit_failure),
    single_update_boost_success = plogis(qlogis(start_p) + learning_strength*success_scale) - start_p,
    single_update_boost_failure = plogis(qlogis(start_p) + learning_strength*failure_scale) - start_p
  )
}

report_pure_paths(0.015)

```

### Expressive noise calibration

On trials where agents are expressive, the produce signal mechanics gets overridden by a direct random draw with a mean = the center of the target semantic attractor, and a sd = expressive_noise_sd. The expressive noise is a function of the circle radius, calibrated to $\frac{\text{circle_radius}}{4.8} = 0.0625$. While this allows for some noise around the signal, the low sd makes an escape from the semantic target center highly improbable, preserving the iconicity of the generated signal on expressive trials.

```{r}
# # what fraction of expressive productions end up far enough from the semantic prototype
# # that they leave its own attractor circle (and so are no longer really iconic)?
#
# estimate_expressive_iconicity_loss <- function(target_center, expressive_noise_sd,
#                                                circle_radius, n_draws = 20000) {
#   x <- pmax(0, pmin(1, rnorm(n_draws, mean = target_center[1], sd = expressive_noise_sd)))
#   y <- pmax(0, pmin(1, rnorm(n_draws, mean = target_center[2], sd = expressive_noise_sd)))
#   dist_to_target <- sqrt((x - target_center[1])^2 + (y - target_center[2])^2)
#   mean(dist_to_target >= circle_radius)
# }
# 
# # Example sweep:
# for (sd in c(0.03, 0.05, 0.0625, 0.08, 0.1)) {
#   loss <- estimate_expressive_iconicity_loss(
#     target_center = c(0.15, 0.85), expressive_noise_sd = sd, circle_radius = 0.3
#   )
#   cat("expressive_noise_sd =", sd, "-> un-iconic rate =", round(loss, 5), "\n")
# }

```

### Iconicity weight

The iconicity recognition bias is governed by two parameters: signal evidence and iconicity weight. Signal evidence is the distance from signal to the center of a target semantic attractor; this evidence is positive for signals inside the correct semantic attractor, and negative if inside the incorrect semantic attractor, thus boosting iconic signals and punishing anti-iconic signals. All output probabilities are clamped to a ceiling of $0.95$ (and a floor of $0.00$) to reflect a baseline lapse rate in recognition. The evidence is further multiplied by iconicity_weight. This parameter is calibrated to 3, which corresponds to a \~0.59 absolute boost for the perfectly iconic signal (evidence = 1), and a \~0.28 absolute punishment for the perfectly anti-iconic signal, when the signal receiver's associative strength = 0.3.

```{r}
# # calibrate iconicity for both perfectly iconic and anti-iconic signals
# test_iconicity_impact <- function(weights = c(0.5, 1, 2, 3, 4), p_refs = c(0.3, 0.5)) {
# 
#   expand_grid(
#     p_ref = p_refs,
#     weight = weights,
#     signal_type = c("Perfectly Iconic (+1)", "Anti-Iconic (-1)")) %>%
#     mutate(
#       icon_ev = if_else(signal_type == "Perfectly Iconic (+1)", 1, -1),
#       logit_guess = qlogis(pmax(1e-9, pmin(1 - 1e-9, p_ref))),
#       new_prob = clamp02(plogis(logit_guess + weight * icon_ev)),
#       # Absolute change in percentage points
#       impact_pp = round((new_prob - p_ref) * 100, 2),
#       # Relative percentage change in probability (% increase / decrease)
#       pct_change_rel = round(((new_prob - p_ref) / p_ref) * 100, 2),
#       new_prob = round(new_prob, 4)) %>%
#     select(p_ref, weight, signal_type, new_prob, impact_pp, pct_change_rel)
# }
# 
# d.iconicity_test <- test_iconicity_impact(
#   weights = c(0.5, 1, 2, 3, 4),
#   p_refs = c(0.3, 0.5))

# note on iconicity assymmetry (for boost/punish iconic/anti-iconic signals)
# Probability shifts are calculated via logit space (qlogis/plogis).
# At baseline p < 0.5 (e.g., p = 0.3):
# 1. Sigmoid slope: +1 shifts logits toward the steepest part of the curve (p = 0.5),
#    whereas -1 pushes into the flatter lower tail (diminishing absolute drops).
# 2. Relative % math: Losses are bounded at -100% (p -> 0), while gains can scale > +200%.
# Note that at p = 0.5, updates are 100% symmetric until hit by the clamp02 ceiling (0.95).

```

## Grid searches through the parameter space

We conducted several grid searches of the parameter space to explore different combination of parameter values and their effect on the resulting iconicity. The initial grid searches explored larger ranges of parameter values but smaller number of simulations per parameter combination (n_sim = 20). The grid searches reported here focused on a smaller, more targeted range of values, but with larger number of simulations per combination (n_sim = 100). Recognition bias and expressive probability were fit as separate grids but shared values for the other parameters. The following parameter values were explored: circle_radius = 0.275, 0.4; drift_sd = 0.125, 0.2, 0.275; center_sd = 0.03, 0.11, 0.24; learning_strength = 0, 0.018, 0.036, 0.054, 0.072, 0.09; iconicity_weight = 0, 0.429, 0.8571, 1.286, 1.714, 2.143, 2.571, 3, 3.429, 3.857, 4.286, 4.714, 5.143, 5.571, 6; expressive_probability = 0.02, 0.04, 0.06, 0.08, 0.1, 0.12, 0.14, 0.16, 0.18, 0.2.

```{r}
# GRID SEARCHES
RESET_MODELS <- FALSE  # set TRUE to force a rerun

plan(multisession, workers = parallel::detectCores() - 1)

N_ROUNDS <- 50

empty_df <- data.frame(
  sim = integer(), gen = integer(), round = integer(), trial = integer(),
  trial_counter = integer(), referent = integer(),
  speaker = character(), listener = character(), type = character(),
  produced_signal = I(list()), dist_to_nearest = numeric(), in_attractor = logical(),
  attractor_id = integer(), is_semantic_attractor = logical(), is_correct_semantic_attractor = logical(),
  old_stored_signal_A = I(list()), new_stored_signal_A = I(list()),
  old_stored_signal_B = I(list()), new_stored_signal_B = I(list()),
  prob = numeric(), success = integer(), evidence = numeric(),
  is_expressive_trial = logical(),
  old_guess_A = numeric(), new_guess_A = numeric(),
  old_guess_B = numeric(), new_guess_B = numeric(),
  stringsAsFactors = FALSE)

# Shared "nuisance" grid, coarse, reused for both mechanisms
d.nuisance_grid <- expand.grid(
  drift_sd          = c(0.125, 0.200, 0.275), #= seq(0.05, 0.35, length.out = 5),
  learning_strength = c(0, 0.018, 0.036, 0.054, 0.072, 0.090), # seq(0, 0.05, length.out = 5),
  circle_radius     = seq(0.15, 0.4, length.out = 3), 
  center_sd_ratio   = c(0.2, 0.4, 0.6)) %>%
  mutate(center_sd = circle_radius * center_sd_ratio)

compute_iconicity <- function(history, n_rounds, cutoff = 0.8) {
  d.iconicity <- history %>%
    mutate(total_round = (generation - 1) * n_rounds + round) %>%
    group_by(simulation, total_round) %>%
    summarise(evidence = mean(evidence, na.rm = TRUE), .groups = "drop")
  
  max_round <- max(d.iconicity$total_round)
  threshold <- max_round * cutoff
  
  d.iconicity %>%
    filter(total_round >= threshold) %>%
    summarise(mean_iconicity = mean(evidence, na.rm = TRUE)) %>%
    pull(mean_iconicity)
}

## GRID 1: recognitionBias
if (RESET_MODELS || !file.exists("models/grid-search-recognitionBias.rds")) {
  
  d.grid.recognitionBias <- d.nuisance_grid %>%
    tidyr::crossing(iconicity_weight = seq(0, 6, length.out = 15)) %>% #seq(0, 0.8, length.out = 10)
    mutate(iconicity = NA_real_, history = vector("list", n()))
  
  for (i in seq_len(nrow(d.grid.recognitionBias))) {
    p <- d.grid.recognitionBias[i, ]
    history <- run_interaction_sim(
      data = empty_df, n_sim = 100, n_referents = 4, n_generations = 1, n_rounds = N_ROUNDS,
      drift_sd = p$drift_sd, learning_strength = p$learning_strength,
      recognition_bias = TRUE, iconicity_weight = p$iconicity_weight,
      circle_radius = p$circle_radius, trap_center_sd = p$center_sd,
      expressive_agents = FALSE,
      success_scale = 7.5, failure_scale = 1
    )
    d.grid.recognitionBias$history[[i]] <- history
    d.grid.recognitionBias$iconicity[i] <- compute_iconicity(history, n_rounds = N_ROUNDS)
    if (i %% 50 == 0) message("recognitionBias: ", i, " / ", nrow(d.grid.recognitionBias))
  }
  saveRDS(d.grid.recognitionBias, "models/grid-search-recognitionBias.rds", compress = TRUE)
} else {
  d.grid.recognitionBias <- readRDS("models/grid-search-recognitionBias.rds")
}

## GRID 2: expressiveAgents
if (RESET_MODELS || !file.exists("models/grid-search-expressiveAgents.rds")) {
  
  d.grid.expressiveAgents <- d.nuisance_grid %>%
    tidyr::crossing(expressive_probability = seq(0.02, 0.20, by = 0.02)) %>%
    mutate(iconicity = NA_real_, history = vector("list", n()))
  
  for (i in seq_len(nrow(d.grid.expressiveAgents))) {
    p <- d.grid.expressiveAgents[i, ]
    history <- run_interaction_sim(
      data = empty_df, n_sim = 20, n_referents = 4, n_generations = 1, n_rounds = N_ROUNDS,
      drift_sd = p$drift_sd, learning_strength = p$learning_strength,
      recognition_bias = FALSE, iconicity_weight = 0,
      circle_radius = p$circle_radius, trap_center_sd = p$center_sd,
      expressive_agents = TRUE, expressive_probability = p$expressive_probability,
      success_scale = 7.5, failure_scale = 1
    )
    d.grid.expressiveAgents$history[[i]] <- history
    d.grid.expressiveAgents$iconicity[i] <- compute_iconicity(history, n_rounds = N_ROUNDS)
    if (i %% 50 == 0) message("expressiveAgents: ", i, " / ", nrow(d.grid.expressiveAgents))
  }
  saveRDS(d.grid.expressiveAgents, "models/grid-search-expressiveAgents.rds", compress = TRUE)
} else {
  d.grid.expressiveAgents <- readRDS("models/grid-search-expressiveAgents.rds")
}

```

```{r plot-grid}

combined_range <- range(c(d.grid.recognitionBias$iconicity, d.grid.expressiveAgents$iconicity), na.rm = TRUE)

# adjust this to try different learning strength (0, .0125, .025...)
ls_fixed <- 0.015
ls_fixed <- unique(d.grid.recognitionBias.new$learning_strength)[5]

plot_grid_faceted <- function(d.grid, x_var, x_lab, fixed_learning_strength, limits) {
  d.grid %>%
    filter(learning_strength == fixed_learning_strength) %>%
    ggplot(aes(x = factor(round(.data[[x_var]], 3)),
               y = factor(round(drift_sd, 3)),
               fill = iconicity)) +
    geom_tile() +
    geom_text(aes(label = round(iconicity, 2)), size = 2.5, color = "black") +
    scale_fill_distiller(palette = "RdBu", direction = -1,
                         limits = limits,
                         rescaler = ~ scales::rescale_mid(.x, mid = 0, to = c(0, 1),
                                                          from = limits)) +
    facet_grid(center_sd_ratio ~ circle_radius, labeller = label_both) +
    #facet_grid(learning_strength ~ center_sd_ratio + circle_radius, labeller = label_both) +
    theme_minimal() +
    labs(x = x_lab, y = "Drift SD", fill = "Iconicity",
         title = paste0("Learning strength = ", round(fixed_learning_strength, 3)))
}

p.recognitionBias <- plot_grid_faceted(d.grid.recognitionBias, "iconicity_weight",
                                       "Iconicity weight", ls_fixed, combined_range)
p.expressiveAgents <- plot_grid_faceted(d.grid.expressiveAgents, "expressive_probability",
                                        "Expressive probability", ls_fixed, combined_range)

# #to derive a radius value for a given center_sd_ratio in the plot, do this:
# d.nuisance_grid %>%
#   filter(center_sd_ratio == 0.2) %>%
#   select(circle_radius, center_sd_ratio, center_sd) %>%
#   distinct() %>%
#   arrange(circle_radius)

# # Compare where the grids peak in iconicity for a given drift_sd
# d.grid.compare <- bind_rows(
#   d.grid.recognitionBias %>%
#     group_by(drift_sd) %>%
#     summarise(peak_iconicity = max(iconicity, na.rm = TRUE), .groups = "drop") %>%
#     mutate(mechanism = "Iconicity recognition bias"),
#   d.grid.expressiveAgents %>%
#     group_by(drift_sd) %>%
#     summarise(peak_iconicity = max(iconicity, na.rm = TRUE), .groups = "drop") %>%
#     mutate(mechanism = "Expressive agents"))
# 
# ggplot(d.grid.compare, 
#        aes(x = drift_sd, 
#            y = peak_iconicity, 
#            color = mechanism)) +
#   geom_line(linewidth = 1) + 
#   geom_point() +
#   scale_y_continuous(limits = c(-1, 1)) + 
#   theme_minimal() +
#   labs(x = "Drift SD", 
#        y = "Peak iconicity (best combo of all other params)", 
#        color = "Mechanism")

```

```{r}
# Metric for evaluating parameter combinations; identify the proportion of stored signals within the iconic category circle 
# in the last 5 interactions
d.grid.eval.Bias <- d.grid.recognitionBias %>%
  select(-center_sd_ratio) %>%
  mutate(
    history = map(history, function(sim_df) {
      sim_df %>%
        group_by(simulation) %>%
        # Keep only the top 5 highest trial_counter values per simulation
        slice_max(order_by = trial_counter, n = 5) %>%
        ungroup()
    }),
    # Calculate the overall proportion of correct signals across those extracted trials
    prop_correct_semantic = map_dbl(history, function(sim_df) {
      mean(sim_df$is_correct_semantic_attractor == TRUE, na.rm = TRUE)
    }))

d.grid.eval.expr <- d.grid.expressiveAgents %>%
  select(-center_sd_ratio) %>%
  mutate(
    history = map(history, function(sim_df) {
      sim_df %>%
        group_by(simulation) %>%
        # Keep only the top 5 highest trial_counter values per simulation
        slice_max(order_by = trial_counter, n = 5) %>%
        ungroup()
    }),
    # Calculate the overall proportion of correct signals across those extracted trials
    prop_correct_semantic = map_dbl(history, function(sim_df) {
      mean(sim_df$is_correct_semantic_attractor == TRUE, na.rm = TRUE)
    }))  

# plotting function for evaluating grids
plot_grid_eval <- function(
    data,
    x_param,
    y_param,
    facet_row = NULL,
    facet_col1 = NULL,
    facet_col2 = NULL,
    prop_col = "prop_correct_semantic",
    title = "Semantic Attractor Convergence Landscape",
    subtitle = "Proportion of last 5 trials in correct semantic attractor"
) {
  
  # identify all mapped parameter columns
  mapped_params <- c(x_param, y_param, facet_row, facet_col1, facet_col2)
  mapped_params <- mapped_params[!is.null(mapped_params)]
  
  # Check if columns exist in dataframe
  missing_cols <- setdiff(c(mapped_params, prop_col), colnames(data))
  if (length(missing_cols) > 0) {
    stop("Column(s) not found in dataframe: ", paste(missing_cols, collapse = ", "))
  }
  
  # group & aggregate (averages over any unmapped background parameters)
  plot_df <- data %>%
    group_by(across(all_of(mapped_params))) %>%
    summarise(
      mean_prop = mean(.data[[prop_col]], na.rm = TRUE),
      .groups = "drop") %>%
    mutate(
      x_var = factor(.data[[x_param]]),
      y_var = factor(.data[[y_param]]))
  
  # base Plot
  p <- ggplot(plot_df, aes(x = x_var, y = y_var, fill = mean_prop)) +
    geom_tile(color = "white", linewidth = 0.4) +
    scale_fill_viridis_c(
      option = "viridis",
      labels = percent_format(accuracy = 1),
      limits = c(0, 1),
      name = "Convergence\nProportion") +
    geom_text(
      aes(
        label = percent(mean_prop, accuracy = 1),
        color = mean_prop > 0.5),
      size = 3,
      fontface = "bold",
      show.legend = FALSE) +
    scale_color_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
    labs(
      title = title,
      subtitle = subtitle,
      x = x_param,
      y = y_param) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      strip.background = element_rect(fill = "grey92", color = NA),
      strip.text = element_text(face = "bold", size = 9),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(size = 9),
      plot.title = element_text(face = "bold"),
      legend.position = "right")
  
  # dynamic Facetting Logic (Supports param3 ~ param4 + param5)
  if (!is.null(facet_row) && !is.null(facet_col1) && !is.null(facet_col2)) {
    # 5-parameter grid: facet_row ~ facet_col1 + facet_col2
    p <- p + facet_grid(
      rows = vars(.data[[facet_row]]),
      cols = vars(.data[[facet_col1]], .data[[facet_col2]]),
      labeller = label_both)
  } else if (!is.null(facet_row) && !is.null(facet_col1)) {
    # 4-parameter grid: facet_row ~ facet_col1
    p <- p + facet_grid(
      rows = vars(.data[[facet_row]]),
      cols = vars(.data[[facet_col1]]),
      labeller = label_both)
  } else if (!is.null(facet_col1)) {
    # 3-parameter grid: ~ facet_col1
    p <- p + facet_wrap(vars(.data[[facet_col1]]), labeller = label_both)
  }
  
  return(p)
}


plot_grid_eval(
  data = d.grid.eval.expr,
  x_param = "expressive_probability",      # Param 1 (X axis)
  y_param = "drift_sd",       # Param 2 (Y axis)
  facet_row = "learning_strength",             # Param 3 (Facet Rows)
  facet_col1 = "center_sd",        # Param 4 (Facet Col 1)
  facet_col2 = "circle_radius"# Param 5 (Facet Col 2)
)

plot_grid_eval(
  data = d.grid.eval.Bias,
  x_param = "iconicity_weight",      # Param 1 (X axis)
  y_param = "drift_sd",       # Param 2 (Y axis)
  facet_row = "learning_strength",             # Param 3 (Facet Rows)
  facet_col1 = "center_sd",        # Param 4 (Facet Col 1)
  facet_col2 = "circle_radius"# Param 5 (Facet Col 2)
)

```
